### 2024-09-04 Week 2 Journal - Grammar of Graphics/week2.qmd
Week 2 initial commit

---
format: 
  pdf:
    linestretch: 2
bibliography: references.bib
---

Tyler Wiederich | STAT 992 | [GitHub Repo](https://github.com/TWiedRW/stat-992)

# Prompt

Read the following introductory material on the grammar of graphics. In your 1-2 page reflection (see rubric for grading criteria), consider the scope of the grammar of graphics compared to your experience with plotting libraries in R (such as base graphics and ggplot2), SAS, and python.

Then, consider the low-level concepts of marks and channels discussed in Ch. 5 of Visualization Analysis and Design. How do these connect in to your understanding of the grammar of graphics?

# Response

The grammar of graphics outlines a process for which visualizations are created.

### 2024-09-30 paper-GoG/paper.qmd
Added paper

---
title: "A Critical Examination of the Grammar of Graphics"
author: Tyler Wiederich
date: today
date-format: long
abstract: |
  This paper explores the philosophy and applications of the Grammar of Graphics [@wilkinson2010]. The Grammar of Graphics uses seven orthogonal classes to deconstruct the process of creating statistical graphics. Through evidence-based reasoning, the classes can be reduced to a minimally sufficient set of features. I will then propose new additions to the minimum set of features for a more robust grammar for creating statistical graphics.
execute: 
  echo: false
  warning: false
columns: 2
format:
  wordcount-pdf:
    fig-align: center
    fig-width: 6
    fig-height: 4
bibliography: references.bib
---

```{=html}
<!-- In this paper, you will examine the grammar of graphics. Your paper should adhere to the following general outline:

-   introduce and summarize the philosophy of the grammar of graphics

-   describe the essential (minimal set of) components of an implementation of the grammar of graphics, based on the theoretical description of the grammar in Wilkinson's book, making an argument for which components of the grammar are essential.

-   evaluate whether VegaLite and ggplot2 contain all of the identified components and compare/contrast the implementation of those components.

-   describe any additions to this minimal set of features/mappings/specifications would you include in a new implementation of the grammar. Justify your inclusions and argue for why they enrich the grammar and facilitate different representations or comparisons of data.

Your paper should contain an introduction describing the grammar and conclude with the additions to the grammar you propose. Empirical claims, ideas, and referenced software should be cited appropriately and specifically (e.g. if talking about an implementation, cite the specific page/section of the implementation's documentation).  -->
```
# Introduction

Data visualizations are an essential method in conveying information [@tukey1965]. When presented with data, these graphics provide faster comprehension, albeit slightly less accurate, than data presented in a table or text format [@meyer1997; @prasad2012; @brewer2012]. The computer boom in the later half of the twentieth century allowed for new innovations in graphics, along with the ability to create graphics faster and more easily [@tukey1965; @beniger1978]. The importance of data visualization is evident from its widespread use in science and media, which means that creating these graphics requires careful considerations to display meaningful information.

However, the mapping of data to the visual space inherently has creative freedom which results in nearly infinite representations. An example of this is produced in @fig-examp1, where the same bivariate data is represented in three different forms. Good practices have developed over the years for producing readable and effective graphics [@tufte1983; @kosslyn1989; @cleveland1984; @heer2010]. A byproduct of graphics research is the recommendations for the forms for which graphics can take. Most modern graphical software packages, such as `ggplot2` [@ggplot2] from R [@Rsoftware], are able to create many types of graphics with minimal user effort. These packages inherently need to follow a well-designed process to produce statistical graphics.

```{r}
#| label: fig-examp1
#| layout-ncol: 3
#| fig-width: 2
#| fig-cap: "Three graphical representations of identical data. The line chart in (a) is useful for showing a trend in y across x. When x is categorical, as in (b), y can be compared across the categories of x. Lastly, the countour plot in (c) shows the bivariate density of x and y. Each graphic conveys information from different contexts of the data."
#| fig-subcap: 
#|  - "Line chart"
#|  - "Bar chart"
#|  - "Contour plot"
library(tidyverse)
set.seed(992)
x = 1:5
y = round(runif(length(x), 5, 15))
theme_set(theme_bw() + theme(aspect.ratio = 1))
p1 <- ggplot(mapping = aes(x = x, y = y)) + scale_y_continuous(limits = c(0, 15))
#p1 + geom_point()
p1 + geom_line()
p1 + geom_bar(stat = 'identity', width = 1, color = 'black', fill = 'skyblue')
p1 + geom_density2d(color = 'skyblue')
```

Since the graphical representation of data can take many forms, it is important to establish a set of rules, or grammar, to define the process of creating data visualizations. One such proposal was the Grammar of Graphics by @wilkinson2010. The Grammar of Graphics defines the process as a sequential set of steps, where each step equally contributes to the rendered graphic. This process includes the following steps: variables, algebra, scales, statistics, geometries, coordinates, and aesthetics. Taken together, these steps provide a system of creating graphics by reducing each step into its most basic form.

In this paper, I will discuss the Grammar of Graphics by @wilkinson2010 and its applications in creating graphics. The grammar, while created with good intentions, has some philosophical flaws that I will address. These shortcomings are both in theory and application of the grammar, from which easy adaptations will allow it to functionally behave with existing graphics tools.

# Grammar of Graphics

@wilkinson2010 describes the Grammar of Graphics as a system of seven orthogonal classes. These classes produce an abstract sequential formulation for creating graphics. Wilkinson states nearly all statistical graphics can be created using the classes, but deviation from the classes or sequence can result in meaningless graphics. The Grammar of Graphics consist of the following seven classes: variables, algebra, scales, statistics, geometries, coordinates, and aesthetics.

### Variables

@wilkinson2010 states an important distinction between the definition of a *variable* and a *varset*. A *variable* is a mapping of objects to values, whereas a *varset* is a mapping of values to all possible ordered lists of objects. This distinction is claimed to simplify the process of algebra, although many graphics packages do not acknowledge the difference. The Graphics Production Language (GPL) is one such example of where the distinction is directly implemented [@wilkinson2005], as well as at `graphic` package for Flutter [@graphic]. In the case of GPL, the distinction allows for data sources to take abstract forms that do not limit graph types to specific data structures [@wilkinson2000].

### Algebra

The algebra class is a set of operations for producing combinations of variables given varsets. This step consists of three operations: cross (\*), nest (/), and blend (+). Cross is the horizontal, or column, concatenation. Nest, while similar to cross, is essentially a left join that creates a nested structure. Blend is the vertical concatenation, like append, that should be used only when the varsets have matching or compatible units. These operations allow for a restructuring of the data, but the ending result resembles tidy data [@JSSv059i10].

### Scales

Scales are transformations on the units of the varsets. These are done before statistical calculations to avoid issues with meaningless results. This is easy to see in the case of averaging logarithmic-transformed data, where given a set of observations $x_i$ for $i=1,\dots,n$

$$\log(\bar{X})=\log(\frac{1}{n}\sum_{i=1}^n x_i)\neq\frac{1}{n}\sum_{i=1}^n \log(x_i)\qquad x_i>0$$

For example, the field of food science tends to use the logarithmic colony-forming units (log CFU) when measuring bacteria counts due to extremely large responses [@johnston2005; @abadias2008]. The interest is in the log CFU units, so averages and other statistical calculations are performed on the log scale. Adjustments to the scales before calculating statistics helps to avoid misrepresented values.

### Statistics

The statistics class takes a varset and outputs a summary varset. Some cases simply use the identity for its statistic, where each observation gets mapped. In other cases, summaries of the varsets can be calculated, such as counts, averages, predicted values from models (such as regression or clustering).

### Geometries

The abstract mapping of data into the visual space is performed with the geometry class. This takes the output of the previous classes and arranges the form that their visual representation will take. This includes the geometric shapes of points, lines/paths, areas, bars/intervals, schemas (boxplots), polygons, contours, and edges. The geometry class produces geometric sets, but still requires further specification before it enters the visual space.

### Coordinates

The coordinate class is how geometries are mapped into the visual space. The most common system is the Cartesian coordinate system, where up to three dimensions can be shown at a time. In most cases, only two dimensions are used since three dimensional representations provide less accurate comparisons [@croxton1932; @cleveland1984; @barfield1989; @fischer2000; @hughes]. Polar coordinates are another common coordinate system when circular geometries are desired, such as pie charts.

```{r}
#| label: fig-examp2
#| layout-ncol: 2
#| fig-width: 1.5
#| fig-height: 3
#| fig-cap: "Two examples of bar charts in both Cartesian and polar coordinates. Figures (a) and (b) are identical with the exception of the coordinate system. Similarly for figures (c) and (d)."
#| fig-subcap: 
#| - "Stacked bar chart in Cartesian coordinates."
#| - "Pie chart in polar coordinates."
#| - "Bar chart in Cartesian coordinates."
#| - "Radial chart in polar coordinates."
p2 <- ggplot(mapping = aes(x = '', y = y, fill = factor(x))) + 
  geom_bar(stat = 'identity', width = 1, position = 'stack') + 
  scale_fill_brewer(palette = 'Set2') + theme_minimal() + 
  theme(legend.position = 'none', aspect.ratio = 1, axis.ticks.x = element_blank(),
                                   panel.grid.major.x = element_blank()) + 
  labs(fill = '', x = '', y ='')
p2
p2 + coord_polar('y') + theme_void() + theme(aspect.ratio = 1, legend.position = 'none')

p3 <- ggplot(mapping = aes(x = x, y = y, fill = factor(x))) + 
  geom_bar(stat = 'identity', width = 1, position = 'stack') + 
  scale_fill_brewer(palette = 'Set2') + theme_minimal() + 
  theme(legend.position = 'none', aspect.ratio = 1, axis.ticks.x = element_blank(), panel.grid.major.x = element_blank()) + 
  labs(fill = '', x = '', y = '')
p3
p3 + coord_polar('y') + theme_void() + theme(aspect.ratio = 1, legend.position = 'none')
```

### Aesthetics

The final class, aesthetics, is the actual mapping of geometries onto the coordinate space. This includes mapping varsets to the coordinates of the geometries and other components, such as color, shape, and size. Aesthetics can also include sound and motion, extending the Grammar of Graphics beyond a static visual space.

## Philosophy

The purpose of the Grammar of Graphics [@wilkinson2010] was to establish a mathematics-based foundation by using a series of mapping functions. Data is initially mapped to varsets before using set operations to join varsets together. These varsets then have direct numeric operations with the scales and statistics. Finally, the varsets are mapped into the visual space with geometries, coordinate systems, and aesthetics. The steps should compliment each other to produce other meaningful statistical graphics.

The ability to create statistical graphics using the Grammar of Graphics does not guarantee that the all permutations of classes produce meaningful graphics. For example, consider @fig-examp-density-polar with the penguins [@pkg-palmerpenguins] dataset. Here, the Cartesian coordinate display of a contour plot is much easier to understand than the same data represented in polar coordinates. While the Grammar of Graphics is capable of producing many types of charts, it does not provide guidance on what charts are useful for users.

```{r}
#| label: fig-examp-density-polar
#| layout-ncol: 2
#| fig-cap: "Contour plots of `penguins` dataset from the palmerpenguins package in R. The left panel is in Cartesian coordinates and the right panel is in polar coordinates. The polar coordinates are more difficult to understand than the Cartesian coordinates when extracting information about the different species."
#| fig-subcap: 
#| - "Cartesian coordinates"
#| - "Polar coordinates"

library(palmerpenguins)
p5 <- ggplot(penguins, mapping = aes(x = bill_length_mm, 
                               y = bill_depth_mm,
                               color = species)) + 
  geom_density2d() + theme(legend.position = 'bottom')
p5
p5 + coord_polar()

```

# Modern Visualization Tools

The classes contained within the Grammar of Graphics [@wilkinson2010] are used in many modern graphics packages [@ggplot2; @Waskom2021; @satyanarayan2017]. These packages typically use similar terms found in the Grammar of Graphics, such as data, scales, geometry, coordinates, and aesthetics. Code examples are provided with @lst-examp3-r and @lst-examp3-python. Using `ggplot2` with R, data is provided in the `ggplot()` function, and aesthetics are mapped using a dedicated `aes()` function. Here, the coordinate positions are included within the aesthetics along with color. Python's `seaborn` module initially calls the geometry with `os.scatterplot()`, using function arguments to identify data, coordinate positions, and color. The plot from @lst-examp3-r is produced in @fig-examp3-r.

```{r}
#| lst-label: lst-examp3-r
#| lst-cap: "Code to produce scatterplot of `penguins` data using `ggplot2` in R."
#| echo: true
#| eval: false

library(palmerpenguins)
data("penguins")
ggplot(data = penguins, mapping = aes(x = bill_length_mm, 
                                      y = bill_depth_mm,
                                      color = species)) + 
  geom_point()

```

```{python}
#| lst-label: lst-examp3-python
#| lst-cap: "Code to produce scatterplot of `penguins` data using `seaborn` in Python."
#| echo: true
#| eval: false
import seaborn as os
os.scatterplot(data=r.penguins, x="bill_length_mm", 
  y="bill_depth_mm",
  hue="species")
```

```{r}
#| label: fig-examp3-r
#| fig-cap: "Scatterplot of `penguins` from @lst-examp3-r using `ggplot2` in R."
#| echo: false
#| eval: true
#| fig-height: 3

library(palmerpenguins)
data("penguins")
ggplot(data = penguins, mapping = aes(x = bill_length_mm, 
                                      y = bill_depth_mm,
                                      color = species)) + 
  geom_point()
```

The Grammar of Graphics was built around GPL, and thus needs adaptations for general use. Consider @lst-examp3-python. To create a graphic, all that was minimally required was data and a geometry mapping to coordinates x and y (hue was not necessary to create the plot). The same can be said for @lst-examp3-r, although the geometry step is more explicit as its own step. A commonality is that both figures were produced with tidy data [@JSSv059i10]. However, most classes from the Grammar of Graphics are either used inherently or by default settings. Therefore, an adaptation of the Grammar of Graphics can be considered by redefining classes and their practicality in modern data visualizations.

## Evaluation of the Seven Classes

Establishing variables from data is an essential step in the Grammar of Graphics. This is obvious from the fact that statistical graphics are the visual mapping of data. The only requirement from the variable class is to create varsets, the mapping of values to all possible ordered lists of objects. From a practical point of view, a variable is simply a $n\times 1$ vector of values that share the same data type.

The algebra step is the first class from the Grammar of Graphics that is implicitly performed by the data structure. For a single data set that is already in a tidy format [@JSSv059i10], the algebra step can be ignored since no crossing, nesting, or blending is needed. However, data often is stored across multiple sources, such as in a relational database [@alma991029835125406387]. The joining of data results in another dataset, which can then become the new data source. Consider @lst-tbl-join1 using the band_members and band_instruments from the `dplyr` package [@dplyr-pkg]. Here, data is stored separately and then joined to create a new data object. Now considering the joined data as the new data source, the algebra step is no longer required. This shows that algebra is only an intermittent step in producing data.

```{r}
#| lst-label: lst-tbl-join1
#| lst-cap: "Full join of `band_members` and `band_instruments` from the `dplyr` package in R. The joined data is stored into `band_info` object."
#| echo: true
#| eval: false
band_info <- dplyr::band_members %>% 
  full_join(dplyr::band_instruments, by = 'name')
```

The next two classes, scales and statistics, follow a similar logic to that of the algebra class. For scales, a transformation onto a variable can be used to create a new dataset with the newly defined variable. Statistics follow a similar process, where summaries and models can be calculated and arranged into a tidy format. @lst-tbl-agg1 shows an example of a scale transformation and averages, where the resulting data is saved into a new data object.

```{r}
#| lst-label: lst-tbl-agg1
#| lst-cap: "Data aggregation using scales and statistics on the `penguins` dataset from the `palmerpenguins` package in R. Data is saved into a new data set called `penguins_summary`."
#| echo: true
#| eval: false

penguins_summary <- penguins %>% 
  mutate(bill_length_cm = bill_length_mm/10,
         bill_depth_cm = bill_depth_mm/10) %>% 
  group_by(species, island, sex, year) %>% 
  summarise(avg_bill_length_cm = mean(bill_length_cm, na.rm = T),
            avg_bill_depth_cm = mean(bill_depth_cm, na.rm = T))
```

Thus far, the classes of algebra, scales, and statistics are argued to be non-essential in the Grammar of Graphics framework. The reasoning is that these classes are able to result in a new dataset that foregoes the necessity to reevaluate the data aggregation steps. In reality, the application of these classes is more fluid than having a strict ordering. For example, the magrittr package [@pkg-magrittr] in R allows for piping, a "this-then-that" approach, where data does not need to be saved into a new object. Additionally, many graphical packages allow for transformations withing the geometry class. Consider @fig-examp4 with the penguins dataset [@pkg-palmerpenguins] where the binning statistic is used within the geom_bar() function. The binning could have taken place before introducing the data into the ggplot() function, but including the transformation with the geometry increases readability and makes it easier to create the graphic. It is not always necessary to calculate relevant transformations and/or statistics as their own steps, so they can be considered non-essential classes of the Grammar of Graphics methodology.

```{r}
#| label: fig-examp4
#| fig-width: 6
#| fig-height: 3
#| fig-cap: "Histogram of bill length (mm) from the `penguin` dataset. Binning is performed inside the geometry of the bar chart."
#| echo: true

ggplot(penguins, mapping = aes(x = bill_length_mm)) + 
  geom_bar(stat = 'bin', fill = 'skyblue', color = 'black') + 
  theme(aspect.ratio = 1/2)
```

The mapping of data into an abstract geometric space includes the classes of geometry and coordinates. The geometry class is essential in defining the visual form of the data, which is the key point of making graphics. However, the placement of geometry requires a coordinate system, which makes the coordinate class inherently essential. Given the limitations of the physical world and computers, displays of dimensional data are restricted to a maximum of three dimensions. The standard Cartesian coordinates have been around since at least the 17th century, making them an ideal default system [@vince2010, chap. 5]. The geometry and coordinate classes work in tandem to create the form that graphic will use in its final product.

The final class is aesthetics. By technical definitions, mapping data to aesthetics is not necessary in producing charts, with the exception of the coordinates of the geometries. @fig-examp4 is one such example where only geometries are mapped to locations. However, it is often useful to encode additional information through one or more aesthetics [@wilkinson2005]. @fig-examp5-allgog shows separate tasks from aesthetics, where only geometry mapping is essential in creating a graphic. Aesthetics are

```{r}
#| label: fig-examp5-allgog
#| fig-cap: "Demonstration of each visual class from the Grammar of Graphics. In (a), a coordinate grid is created based on the identify statistic of bill length (mm) and bill depth (mm). Panel (b) uses the point geometry to map the variable pairs onto the coordinate space. Panel (d) map species to color and sex to shape. Only (b) was required to create a graphic from data, but (c) adds additional context."
#| fig-subcap: 
#| - "Cartesian coordinates"
#| - "Aesthetic: coordinate locations"
#| - "Aesthetic: color + shape "
#| layout: [[1,1],[1]]
#| fig-height: 2

p4 <- ggplot(data = penguins, mapping = aes(x = bill_length_mm, y = bill_depth_mm))
p4 
p42 <- p4 + geom_point()
p42
p44 <- p4 + geom_point(aes(color = species, shape = sex)) + theme(aspect.ratio = 1/2, legend.position = 'bottom')
p44

```

## Two Applications of Essential Classes

<!--# -   evaluate whether VegaLite and ggplot2 contain all of the identified components and compare/contrast the implementation of those components. -->

In this section, I will evaluate the graphics software of ggplot2 in R [@wickham2010] and VegaLite in JavaScript [@satyanarayan2017]. The ggplot2 package was created from the Grammar of Graphics philosophy, using a layering approach navigate through the classes. VegaLite was similarly created using the Grammar of Graphics, but opted to use differing terminology and incorporate interactive elements. Each graphics package has its own implementation of the Grammar of Graphics and the minimally required classes.

### ggplot2

-   Example with minimal classes

-

### VegaLite

VegaLite uses *units* to define graphics, which consists of three required classes: data, mark-types, and encodings. Data follows the same form as tidy data [@JSSv059i10] and is clearly an essential component required from the Grammar of Graphics. Mark-types define the geometry of the data before using the visual space. Lastly, encodings are the visual space of geometry location and other features, such as color and shape.

### Comparing ggplot2 and VegaLite

-   Main separating factor is layers vs. JSON

-   Terminology

-   Extent (static vs. interactive)

# A New Grammar

<!--# -   describe any additions to this minimal set of features/mappings/specifications would you include in a new implementation of the grammar. Justify your inclusions and argue for why they enrich the grammar and facilitate different representations or comparisons of data. -->

-   Algebra, scales, and statistics all can form new data

-   Example with exp(normal) data, use model to get data from model before plotting

-   Geometry, coordinates, and aesthetics all used together

-   Need to have readable graph, so use

# Conclusion

\newpage

### References

::: {#refs}
:::

### 2024-10-01 paper-GoG/paper.qmd
Updated paper

@@
 
--   describe the essential (minimal set of) components of an implementation of the grammar of graphics, based on the theoretical description of the grammar in Wilkinson's book, making an argument for which components of the grammar are essential.
+-   describe the essential (minimal set of) components of an implementatiorn of the grammar of graphics, based on the theoretical description of the grammar in Wilkinson's book, making an argument for which components of the grammar are essential.
 
@@
 The computer boom in the later half of the twentieth century allowed for new innovations in graphics, along with the ability to create graphics faster and more easily [@tukey1965; @beniger1978].
-The importance of data visualization is evident from its widespread use in science and media, which means that creating these graphics requires careful considerations to display meaningful information.
+The importance of data visualization is evident from its widespread use in science and media, which means that creating these graphics requires careful consideration to display meaningful information.
 
@@
 @fig-examp5-allgog shows separate tasks from aesthetics, where only geometry mapping is essential in creating a graphic.
-Aesthetics are
+Aesthetics are a semi-required class, noting that only some aesthetic mappings are necessary to produce a visual graphic.
 
@@
 
+To summarize, the only required classes are variables, geometries, coordinates, and aesthetics.
+Variables are the link to the data.
+Geometry is the representation of the variables and is mapped into a space using a coordinate system.
+Lastly, aesthetics bring the graph into the visual space.
+These features form the set of the minimally required classes to produce statistical graphics.
+
 ## Two Applications of Essential Classes
@@
 
-In this section, I will evaluate the graphics software of ggplot2 in R [@wickham2010] and VegaLite in JavaScript [@satyanarayan2017].
+In this section, I will evaluate the grammar of ggplot2 in R [@wickham2010] and VegaLite in JavaScript [@satyanarayan2017].
 The ggplot2 package was created from the Grammar of Graphics philosophy, using a layering approach navigate through the classes.
@@
 
--   Example with minimal classes
+Many of the examples in this paper were created with `ggplot2` [@ggplot2], a package inspired by the Grammar of Graphics.
+To implement the required classes, data is first supplied to the data argument in the `ggplot()` function.
+Then, aesthetics are supplied into the `mapping` argument through the `aes()` function.
+Lastly, a geometry is chosen by adding a `geom_function()` function.
+Alternatively, variables can be directly supplied into the `aes()` function and/or within the `geom_function()` function.
+An example of the minimally required features is supplied in @lst-ggplot-outline.
 
--
+```{r}
+#| lst-label: lst-ggplot-outline
+#| lst-cap: "`ggplot2` code demonostrating the minimum required classes from the Grammar of Graphics. "
+#| eval: false
+#| echo: true
+
+ggplot(data = data, mapping = aes(x = x, y = y)) + 
+  geom_function()
+```
+
+When making graphics with ggplot2, the order is important to consider.
+The evaluation of steps is in the presented order of layers.
+For example, two geometries can be mapped onto the coordinate space, but the latter geometry will overlap the former (@fig-examp6).
+This shows that `ggplot2` uses looped sequence through the Grammar of Graphics, where graphics are layered sequentially after each evaluation of the grammar.
+
+```{r}
+#| label: fig-examp6
+#| fig-cap: "Two scatterplots with different orders of layering `geom_point()` objects.
+When the red points are layered second (left panel), they appear on top of the black points.
+When the red points are layered first (second panel), they disappear behind the black points."
+#| fig-subcap: 
+#| - "Black points mapped before red points."
+#| - "Red points mapped before black points."
+#| fig-height: 2
+#| fig-width: 2
+#| layout-ncol: 2
+ggplot(mapping = aes(x = x, y = y)) + 
+  geom_point(size = 4, color = 'black') + 
+  geom_point(size = 2, color = 'red') 
+ggplot(mapping = aes(x = x, y = y)) + 
+  geom_point(size = 2, color = 'red') + 
+  geom_point(size = 4, color = 'black') 
+```
 
@@
 
-VegaLite uses *units* to define graphics, which consists of three required classes: data, mark-types, and encodings.
-Data follows the same form as tidy data [@JSSv059i10] and is clearly an essential component required from the Grammar of Graphics.
+VegaLite [@satyanarayan2017] uses *units* to define graphics, which consists of three required classes: data, mark-types, and encodings.
+Data follows the same form as tidy data [@JSSv059i10].
 Mark-types define the geometry of the data before using the visual space.
@@
 
+The method of implementing VegaLite is through a JSON syntax.
+Each class is supplied into its corresponding JSON field, which are called *data*, *mark*, and *encoding*.
+In the data field, the file path and format type need to be specified.
+The mark field specifies the geometry.
+Within the encoding field, the x and y coordinates are supplied as their own fields, along with the variable and variable type.
+Overlapping geometries are created by duplicating the individual grammars as objects supplied to a parent `layers` field.
+
 ### Comparing ggplot2 and VegaLite
 
--   Main separating factor is layers vs. JSON
+The primary factor separating `ggplot2` and VegaLite is the syntax.
+`ggplot2` uses layers, which are added onto previous steps.
+Arguments supplied to the `ggplot()` function are applied globally and followed by the addition of extra layers.
+On the contrary, VegaLite compiles its layers through the JSON syntax.
+The JSON syntax only allows for one geometry to be applied at a time, with the exception of geometries created under different encoding
 
--   Terminology
+Another difference between ggplot2 and VegaLite is the terminology.
+`ggplot2` mainly uses the same terminology as the Grammar of Graphics [@wilkinson2010], whereas VegaLite tends to copy the terminology from @heer2012.
+What VegaLite calls "mark" and "encoding" are equivalent to `ggplot2`'s "geom(etry)" and "aes(thethic)", respectively.
+The difference in terminology between these two graphics packages is mostly semantic.
 
--   Extent (static vs. interactive)
+Although interactivity is not considered in the Grammar of Graphics [@wilkinson2010], VegaLite incorporates additional grammar for *selection*.
+Selection consists of the following components: name, type, predicate, domain\|range, event, init, transforms, and resolve.
+This grammar class allows users to interact with values in the graph to create new graphics.
+The default `ggplot2` package does not support *selection*, but the `Plotly` package [@pkg-plotly] is an extension that allows `ggplot2` users to interact with graphics.
 
@@
 
+We have seen that data, geometry, coordinates, and aesthetics are all essential classes to the Grammar of Graphics [@wilkinson2010].
+Without these classes, it would not be possible create graphics.
+However, these essential classes only allow for basic visualizations to be created.
+In this section, I propose a new grammar that adapts the Grammar of Graphics to a practical setting in the modern era.
+
+The first class in my proposed grammar is *data summary*.
+All statistical graphics start from a realization of data, where the data can take many structural forms.
+Computer software for creating graphics mostly rely on tidy data [@JSSv059i10], which is where each row is an observation and each column is a variable.
+What the *data summary* class does is supply a data set after combining variables, transforming scales, and calculating statistics.
+However, some data sources cannot be coerced into a tidy format [@filonik2019], which can be mitigated by using the same aggregation steps to the data strucutre.
+By replacing the first four Grammar of Graphics classes with a unified *data summary* class, the data source is prepared for both graphics and other tasks, like tables or output files.
+The *data summary* class simplifies the grammar by assuming the data is already prepared for graphing.
+
+My next proposed class is *visual mapping*, a class that combines geometries, coordinates, and aesthetics.
+Graphics software generally combine geometries and aesthetics into a single step.
+For example, `ggplot2` can accomplish the combined classes with `geom_function(mapping = aes(x = x, y = y, color = color_var, shape = shape_var, ...)`.
+The distinction between geometry and aesthetics is primarily performed behind the scenes when rendering the graphic, and thus not truly differentiated for the user.
+Graphics software also tends to default to Cartesian coordinates, resulting in alternative coordinate themes only being used when necessary.
+Combining geometry, coordinates, and aesthetics into a unified *visual mapping* class makes it so that the user only considers the components necessary to produce their desired graph.
+
+Before introducing new classes, there is an unmentioned benefit to simplifying the classes into *data summary* and *visual mapping*.
+The strict sequencing of the Grammar of Graphics made it difficult to overlay multiple geometries.
+This was evident with VegaLite [@satyanarayan2017], where each iteration of the grammar was used as its own instance within "layers".
+Combining the classes for each stage into data and visual mappings allows for complicated graphics to be produced without adding further restrictions.
+
+The final class I propose is *graph readability*, a new class that evaluates the effectiveness of statistical graphics.
+This class is designed to ensure that the graph is readable given the nature of the data.
+Similarly to how statistical models can produce nonsensical results if not used correctly, the effectiveness of a graph sometimes hinges on the nature of the data.
+In addition to the structure of the data, research shows that some graphs will outperform other graphs simply by changing the comparison method [@cleveland1984; @heerCrowdsourcingGraphicalPerception2010; @fischer2005].
+Therefore, creating graphics is not the end result and using *graph readability* will help in a finalized product.
+
+Consider @fig-examp7 with zero-inflated Poisson count data.
+When the proportion of zeros is high, including the zeros in the histogram results in difficulties differentiating values for non-zero counts.
+One possible solution is summarize the proportion of zeros as text.
+Another instance shown in @fig-examp8 where a graph is made with points.
+The overlapping points hide the distributions, but adding a small amount of random noise to the placement and adjusting the opacity helps to make the graph more readable.
+In both examples, the graph are constructed "correctly" from a Grammar of Graphics framework [@wilkinson2010].
+However, taking the extra step to in the final design creates more meaningful graphics.
+
+```{r}
+#| label: fig-examp7
+#| fig-cap: "Two histograms of zero-inflated Poisson count data. In the right panel, the percentage of zeros is reported in the caption so that the distribution of counts "
+#| fig-subcap: 
+#| - "Zero counts included in histogram."
+#| - "Zero counts summarized in caption."
+#| layout-ncol: 2
+#| fig-height: 3
+
+n <- 5000
+n_zi <- 4500
+x_zip = c(rep(0, n_zi), rpois(n=n-n_zi, lambda = 10))
+percent_zero <- mean(x_zip==0)*100
+ggplot(mapping = aes(x = x_zip)) + 
+  geom_histogram(binwidth = 1, color = 'black', fill = 'skyblue') + xlab('x')
+ggplot(mapping = aes(x = x_zip[x_zip > 0])) + 
+  geom_histogram(binwidth = 1, color = 'black', fill = 'skyblue') + xlab('x') + 
+  scale_x_continuous() + 
+  labs(caption = paste0(round(percent_zero, 2), '% of counts reported as zero'))
+```
+
+```{r}
+#| label: fig-examp8
+#| fig-cap: "Two plots using points to measure bill length (mm).
+The left panel does not make any adjustments to point locations, resulting in overlap and difficulties determining the distribution.
+The right panel adds a small amount of horizontal jitter and adjusts the opacity of the points to help distinguish the density of the points."
+#| fig-subcap: 
+#| - "Only using `geom_point()`"
+#| - "Using `geom_point()` with a horizontal jitter adjustment."
+#| layout-ncol: 2
+#| fig-height: 3
+
+p8 <- ggplot(penguins, mapping = aes(x = sex, y = bill_length_mm))
+p8 + geom_point()
+p8 + geom_point(position = position_jitter(width = .1), alpha = 1/5)
+```
+
+I will now demonstrate my proposed grammar through an example.
+Consider the `starwars` dataset in the `dplyr` package [@pkg-dplyr].
+Suppose that we want to test the hypothesis that the sex of characters in Star Wars has an association on height.
+A simple analysis of variance model can help to answer the hypothesis, from which the `emmeans` package [@pkg-emmeans] can extract point estimates and confidence intervals.
+This concludes the data summary class.
+The next step is visual mappings, where points and error bars make the sense for plotting estimated heights and their 95% confidence intervals.
+By default assumption, the geometries are positioned on a Cartesian coordinate system.
+Lastly, graph readability is addressed by adjusting they height axis range to include zero so that the scale of the range is not compressed.
+The end result is a decent graph that only considered data, visual mappings, and readability.
+
+```{r}
+library(emmeans)
+starwars <- dplyr::starwars
+model <- aov(height~sex, data = dplyr::starwars)
+em <- emmeans(model, ~sex, infer = c(T,T)) %>% as.data.frame()
+ggplot(em, mapping = aes(x = sex)) + 
+  geom_point(aes(y = emmean)) + 
+  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
+                width = .5) + 
+  scale_y_continuous(limits = c(0, max(em$upper.CL + 25))) + 
+  labs(title = 'Estimated height of Star Wars characters',
+       x = 'Sex', y = 'Predicted height (cm)',
+       caption = '95% confidence intervals')
+```
+
 # Conclusion
+
+Throughout this paper, I discussed the Grammar of Graphics [@wilkinson2010] and its application for creating data visualizations.
+The Grammar of Graphics developed a mathematical framework for creating graphics, where the final result is produced through the final aesthetic class.
+While the framework technically works in creating visualizations, the practical application needs adaptations.
+
+I proposed a new grammar that consists of three classes: data summary, visual mapping, and graph readability.
+These classes combined the classes presented by Wilkinson into a sequence of data, visuals, and proofreading.
+This new grammar allowed for a singular sequence that is
 

### 2024-10-01 paper-GoG/paper.qmd
Final commit

@@
 I will then propose new additions to the minimum set of features for a more robust grammar for creating statistical graphics.
+My new grammar simplifies the Grammar of Graphics into three class focused on data summaries, visual mappings, and graph readability.
 execute: 
@@
 These shortcomings are both in theory and application of the grammar, from which easy adaptations will allow it to functionally behave with existing graphics tools.
+Code for this paper can be found on my GitHub [page](<https://github.com/TWiedRW/stat-992/tree/main>).
 
@@
 This distinction is claimed to simplify the process of algebra, although many graphics packages do not acknowledge the difference.
-The Graphics Production Language (GPL) is one such example of where the distinction is directly implemented [@wilkinson2005], as well as at `graphic` package for Flutter [@graphic].
+The Graphics Production Language (GPL) is one such example of where the distinction is directly implemented [@wilkinson2005], as well as the `graphic` package for Flutter [@graphic].
 In the case of GPL, the distinction allows for data sources to take abstract forms that do not limit graph types to specific data structures [@wilkinson2000].
@@
 Finally, the varsets are mapped into the visual space with geometries, coordinate systems, and aesthetics.
-The steps should compliment each other to produce other meaningful statistical graphics.
+The steps should complement each other to produce other meaningful statistical graphics.
 
-The ability to create statistical graphics using the Grammar of Graphics does not guarantee that the all permutations of classes produce meaningful graphics.
+The ability to create statistical graphics using the Grammar of Graphics does not guarantee that all permutations of classes produce meaningful graphics.
 For example, consider @fig-examp-density-polar with the penguins [@pkg-palmerpenguins] dataset.
@@
 Geometry is the representation of the variables and is mapped into a space using a coordinate system.
-Lastly, aesthetics bring the graph into the visual space.
+Lastly, aesthetics brings the graph into the visual space.
 These features form the set of the minimally required classes to produce statistical graphics.
@@
 
--   Algebra, scales, and statistics all can form new data
-
--   Example with exp(normal) data, use model to get data from model before plotting
-
--   Geometry, coordinates, and aesthetics all used together
-
--   Need to have readable graph, so use
-
 We have seen that data, geometry, coordinates, and aesthetics are all essential classes to the Grammar of Graphics [@wilkinson2010].
-Without these classes, it would not be possible create graphics.
+Without these classes, it would not be possible to create graphics.
 However, these essential classes only allow for basic visualizations to be created.
@@
 What the *data summary* class does is supply a data set after combining variables, transforming scales, and calculating statistics.
-However, some data sources cannot be coerced into a tidy format [@filonik2019], which can be mitigated by using the same aggregation steps to the data strucutre.
+However, some data sources cannot be coerced into a tidy format [@filonik2019], which can be mitigated by using the same aggregation steps to the data structure.
 By replacing the first four Grammar of Graphics classes with a unified *data summary* class, the data source is prepared for both graphics and other tasks, like tables or output files.
@@
 Similarly to how statistical models can produce nonsensical results if not used correctly, the effectiveness of a graph sometimes hinges on the nature of the data.
-In addition to the structure of the data, research shows that some graphs will outperform other graphs simply by changing the comparison method [@cleveland1984; @heerCrowdsourcingGraphicalPerception2010; @fischer2005].
+In addition to the structure of the data, research shows that some graphs will outperform other graphs simply by changing the comparison method [@cleveland1984; @heer2010; @fischer2005].
 Therefore, creating graphics is not the end result and using *graph readability* will help in a finalized product.
@@
 When the proportion of zeros is high, including the zeros in the histogram results in difficulties differentiating values for non-zero counts.
-One possible solution is summarize the proportion of zeros as text.
+One possible solution is to summarize the proportion of zeros as text.
 Another instance shown in @fig-examp8 where a graph is made with points.
 The overlapping points hide the distributions, but adding a small amount of random noise to the placement and adjusting the opacity helps to make the graph more readable.
-In both examples, the graph are constructed "correctly" from a Grammar of Graphics framework [@wilkinson2010].
-However, taking the extra step to in the final design creates more meaningful graphics.
+In both examples, the graphs are constructed "correctly" from a Grammar of Graphics framework [@wilkinson2010].
+However, taking the extra step to proofread the final design creates more meaningful graphics.
 
@@
 
+The Grammar of Graphics did not directly correspond to `ggplot2` or VegaLite.
+These packages both required adaptations from Wilkinson and, in the case of VegaLite, introduced new elements into the grammar.
+While the Grammar of Graphics provided a framework, it was used as a guide for creating new grammars.
+
 I proposed a new grammar that consists of three classes: data summary, visual mapping, and graph readability.
 These classes combined the classes presented by Wilkinson into a sequence of data, visuals, and proofreading.
-This new grammar allowed for a singular sequence that is
+This new grammar allowed for a singular sequence that is focused on application and simplicity.
+Data summary allowed for data aggregation to take place before introducing raw data into the sequence of creating graphics.
+Visual mapping considers all necessary geometries, coordinate systems, and aesthetics simultaneously
 

### 2024-10-01 paper-GoG/paper.qmd
Final corrections

@@
 I will then propose new additions to the minimum set of features for a more robust grammar for creating statistical graphics.
-My new grammar simplifies the Grammar of Graphics into three class focused on data summaries, visual mappings, and graph readability.
+My new grammar simplifies the Grammar of Graphics into three classes focused on data summaries, visual mappings, and graph readability.
 execute: 
@@
 Data summary allowed for data aggregation to take place before introducing raw data into the sequence of creating graphics.
-Visual mapping considers all necessary geometries, coordinate systems, and aesthetics simultaneously
+Visual mapping considers all necessary geometries, coordinate systems, and aesthetics simultaneously.
+Lastly, graphs are proofread in the graph readability class so that they provide meaningful interpretations.
+These three classes simplified the complexity of Wilkinson's grammar and provide a robust method for the applications of creating graphics.
 

### 2024-12-03 paper-vis-inference/paper.qmd
Paper 2 (computer issues?)

---
title: "Paper 2 Title"
subtitle: "A comparative review"
author: Tyler Wiederich
format: 
  wordcount-pdf: 
    wordcount-banner: true
bibliography: references.bib
biblio-style: https://www.zotero.org/styles/apa
---

```{=html}
<!--
In this paper, you will compare visual inference with Bayesian and Frequentist inference techniques such as hypothesis testing and Bayes factors. You should describe the necessary prerequisites for each method (e.g. you cannot do visual inference without some sort of data, null data generation method, and display method) and what conclusions can be drawn from each procedure. Your paper should at least touch on the underlying philosophical and logical approach to each type of inference (see resources list below), as well as the available test statistics and evaluation methods. If you wish, you may argue that one or more inference methods are superior, but it is also sufficient to compare and contrast the inference methods and identify in which situations one is superior.
-->
```

NOTES

-   This paper is about visual inferences versus Frequentist/Bayesian inference, not visual inference for Bayes/Frequentist

```{r}
set.seed(2024)
library(tidyverse)
```

# Introduction

The overall goal of statistics is to quantify uncertainty.

Implicitly, many statistical methods revolve around the concept of a hypothesis test.

However, there is no single approach to statistical inference. (George Box quote?)

Early statistical inference is generally regarded as Frequentist.

In the later 20th century, a newer framework developed on the idea that previously known information can help guide inference, known as Bayesian.

In this paper, I will talk about three types of statistical inferences: Frequentist, Bayesian, and visual. The paper is organized

# Types of Inference

Structure

-   Basic idea

-   Creating a null hypothesis

-   Hypothesis testing

-   Other considerations

-   Issues?

## Frequentist {#sec-frequentist}

The classical framework of statistics revolves around the idea that independent and repeated events follow a stable probability function [@lehmann2008].

```{r}
#| label: fig-rpt-samples
#| fig-cap: "Estimated probability from a Bernoulli distribution with independent and identical samples. As the number of samples increases, the cumulative average estimate for the proportion of success stabilizes around the true probability of 0.7."

n = 1000
tibble(
  n = 1:n,
  x = rbinom(n, size = 1, prob = 0.7)) %>% 
  mutate(p_hat = cummean(x)) %>% 
  ggplot(mapping = aes(x = n, y = p_hat)) + 
  geom_hline(yintercept = 0.7, color = 'red', linetype='dashed') + 
  geom_line() + 
  scale_y_continuous(limits = c(0,1)) + 
  labs(x = 'Repeated Samples',
       y = 'Estimated Probability')

```

In the early development of Frequentist statistics, philosophical debates between R.A. Fisher and J. Neyman led to

Ronald Fisher believed that probability was deductive in nature and that probabilistic events could be formulated mathematically to produce a maximum likelihood function [@fisher1935]. For example, the binomial distribution is given by $P(X=x)={n\choose x}p^x(1-p)^{n-x}$, where $X$ is the random variable for the number of successes out of $n$ trials given a probability $p$ for $x$ successes. The probability mass function comes from multiplying the number of combinations of a series of successes to the probabilities of individual successes and individual failures.

An inductive approach was developed by Jerzy Neyman and Egon Pearson

The Frequentist analysis technique begins with an underlying assumption of the data generating function. In practice, this function is rarely known, and thus is approximated by a reasonable probability function. Some examples include a Poisson distribution for count data, a binomial distribution for proportions of successes, and a normal distribution for unimodal continuous observations.

After data is collected, a hypothesis test can be formulated. The general format of a Frequentist hypothesis test is to assume that a function of the data arose from an assumed data generating function. If the function of the data does not

## Bayesian {#sec-bayesian}

Bayesian inference begins with some knowledge about the parameter of interest.

## Visual {#sec-visual}

Unlike Frequentist and Bayesian inference, visual inference relies on the nuances of human perception. With visual inference, data is presented in the visual space and treated as a statistic. For example, a common assumption for linear models in the form of $Y=X\beta+\epsilon$ is that the residual term, $\epsilon$, is independently and identically distributed with respect to a Normal distribution with a mean of zero and an unknown variance. Formal tests, like the Shapiro-Wilk test \[, can be used to

The null data generation process

One potential issue with visual inference is the human nature of wanting to find patterns in randomness.

# Example {#sec-example}

To effectively demonstrate the three inference types, consider the following example.

# Discussion {#sec-discussion}

# Conclusion {#sec-conclusion}

@gelman2013

### 2025-03-02 paper-vis-inference/paper.qmd
Updates with code and lots of other stuff

@@
 ---
-title: "Paper 2 Title"
-subtitle: "A comparative review"
+title: "Statistical Inference"
+subtitle: "Frequentist, Bayesian, and Visual oh my"
 author: Tyler Wiederich
+date: "`r format(Sys.time(), '%B %d, %Y')`"
 format: 
-  wordcount-pdf: 
-    wordcount-banner: true
+  wordcount-pdf: default
 bibliography: references.bib
 biblio-style: https://www.zotero.org/styles/apa
+execute: 
+  echo: false
+  message: false
+  warning: false
 ---
@@
 
-NOTES
-
--   This paper is about visual inferences versus Frequentist/Bayesian inference, not visual inference for Bayes/Frequentist
-
 ```{r}
@@
 library(tidyverse)
+library(emmeans)
+library(brms)
+theme_set(theme_bw())
 ```
@@
 
-The overall goal of statistics is to quantify uncertainty.
+The quantification of uncertainty is one of the key components behind statistical methodology [@acree2021; @lehman2008book] .
+This is useful in cases where population parameters are unknown or unable to be reasonably measured.
+For example, some possible research questions that would require statistical methods include “What percentage of Minnesotan residents agree with the proposed legislation?,” or “Does the placement of higher margin products on the middle shelf of a store aisle lead to more profits than placement on top and bottom shelves?” These types of questions can be approximated using statistical inference, although the approach differs by the specific method of inference.
 
-Implicitly, many statistical methods revolve around the concept of a hypothesis test.
+Implicitly, many statistical methods revolve around the concept of a hypothesis test [@neyman1933; @acree2021].
+These tests are designed to determine whether or not there is evidence to support a claim.
+Data is evaluated against a null hypothesis, which is typically a baseline assumption or a specific condition.
+If the condition is deemed unlikely to occur by chance, then there is evidence to support its complement, called the alternative hypothesis.
+The strength of evidence can be quantified, leading to decisions about what interpretation or decision to make regarding the results of the method.
 
-However, there is no single approach to statistical inference. (George Box quote?)
+Early statistical methodology is generally regarded as Frequentist inference [@neyman1933; @lehmann2008book].
+The idea behind Frequentist methods is that data is random variable, generated from some probability function using at least one unknown population parameter.
+Functions of the data yield test statistics that are used to evaluate a hypothesis test.
+In Frequentist testing, the null hypothesis is usually defined such that there is no effect, and the alternative hypothesis is that there is an effect.
+Since the data is considered random, statements are made in terms of expected results under repeated sampling.
 
-Early statistical inference is generally regarded as Frequentist.
+A newer framework developed in the later 20th century using the idea that previously known information can help guide inference, which became known as Bayesian inference [@gelman2013; @lehmann2008book].
+The availability of previous information, or lack thereof, forms the basis of belief for what occurs in nature.
+Bayesian inference combines new data with prior knowledge to update the belief of population parameters.
+In this case, the population parameter(s) is considered random and distributed according to belief and observed data.
 
-In the later 20th century, a newer framework developed on the idea that previously known information can help guide inference, known as Bayesian.
+The use of visualizations is a valuable method for understanding the structure and patterns in data [@tukeyEDA], but they can also guide formal analyses [@loy2015; @loy2016; @loy2017].
+Visual inference is where hypotheses can be tested by human perception instead of mathematical formulation.
+For example, Q-Q plots are a common diagnostic tool for checking normality assumptions [@loy2016].
+While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist.
 
-In this paper, I will talk about three types of statistical inferences: Frequentist, Bayesian, and visual. The paper is organized
+This paper is organized as follows. The foundations and uses of Frequentist, Bayesian, and visual inferences are organized XXXXXX
 
@@
 
-Structure
+The role of the population parameter is a key differentiation between Frequentist and Bayesian methodologies [@pek].
+The former assumes the parameter to be a fixed value for its target population [@neyman1977], whereas the latter assumes the parameter is a random variable drawn from a probability distribution [@gelman2013].
+This changes the scope of inference for Frequentist and Bayesian inference.
+On the other hand, visual inference relies on human perception, which can be used on its own or in conjunction with Frequentist and Bayesian methods.
 
--   Basic idea
+## Frequentist Inference {#sec-frequentist}
 
--   Creating a null hypothesis
+The classical framework of statistics revolves around the idea that independent and repeated events follow a stable probability function with a fixed parameter [@lehmann2008].
+As such, many instances of Frequentist inference include some mention of independent and identically distributed random variables.
+In practice, this assumption can be addressed through randomization [@neyman1977] or checking model diagnostics [@loy2016].
 
--   Hypothesis testing
-
--   Other considerations
-
--   Issues?
-
-## Frequentist {#sec-frequentist}
-
-The classical framework of statistics revolves around the idea that independent and repeated events follow a stable probability function [@lehmann2008].
+Frequentists treat population parameters as fixed and unknown values, where data are random variables drawn from
 
@@
   scale_y_continuous(limits = c(0,1)) + 
-  labs(x = 'Repeated Samples',
+  labs(x = 'Number of samples',
        y = 'Estimated Probability')
@@
 
-In the early development of Frequentist statistics, philosophical debates between R.A. Fisher and J. Neyman led to
+In the early development of Frequentist statistics, philosophical debates between R.
+A.
+Fisher and J.
+Neyman led to the distinction between "inductive inference" and "inductive behavior" [@lehmann2008].
+Inductive inferences, as preferred by Fisher, involves logical rational to develop models.
+On the other hand, Neyman's inductive behavior addresses the additional component of randomization through hypothesis testing.
+Both arguments start with the assumption that the random variable in question is independent and identically distributed.
+However, the two statisticians differed on the interpretations and approaches, with Fisher favoring intuition and Neyman favoring proofs.
+The methodology was similar between them, emphasizing specific approaches and scope of inference.
 
-Ronald Fisher believed that probability was deductive in nature and that probabilistic events could be formulated mathematically to produce a maximum likelihood function [@fisher1935].
+R.
+A.
+Fisher believed that probability was deductive in nature and that probabilistic events could be formulated mathematically to produce a maximum likelihood function [@fisher1935].
 For example, the binomial distribution is given by $P(X=x)={n\choose x}p^x(1-p)^{n-x}$, where $X$ is the random variable for the number of successes out of $n$ trials given a probability $p$ for $x$ successes.
-The probability mass function comes from multiplying the number of combinations of a series of successes to the probabilities of individual successes and individual failures.
+The probability mass function comes from multiplying the number of combinations that a series of successes can take to the probabilities of individual successes and individual failures, a formulation of combinatorics.
+When taking the likelihood, the most plausible estimate of the probability can be found by establishing which value maximizes the likelihood with respect to all possible probabilities [@fig-freq-binom].
 
-An inductive approach was developed by Jerzy Neyman and Egon Pearson
+```{r}
+#| label: fig-freq-binom
+#| fig-cap: "Binomial example of R.A.
+Fisher's inductive inference.
+A single observation from a Binomial distribution is drawn and observed to have 13 successes out of 20 trials, where the true probability of success is 0.7.
+The likelihood function is optimized at $p=0.65$ and the red line in (b) shows the true probability."
+#| fig-subcap: 
+#| - "Binomial$(n=20,p=0.7)$"
+#| - "Likelihood of $p$ for 13 successes out of 20 trials."
+#| layout-ncol: 2
+#| out-width: 90%
+
+set.seed(1999)
+n=20
+p=.7
+y=rbinom(1, n, p)
+
+tibble(
+  x = 0:n
+) %>% 
+  mutate(fx = dbinom(x, n, p)) %>% 
+  ggplot(mapping = aes(x = x, y = fx)) + 
+  geom_bar(stat = 'identity', width = 1,
+           color = 'black', fill = 'skyblue') + 
+  geom_text(mapping = aes(x = y, y=0.1), label = 'Observed data', angle = 90,
+            color = 'grey20', hjust=1) + 
+  labs(x = 'Number of successes',
+       y = 'Probability')
+
+tibble(
+  p = seq(0, 1, l=1000)
+) %>% 
+  mutate(L = choose(n,y)*p^(y)*(1-p)^(n-y)) %>% 
+  ggplot(mapping = aes(x = p, y = L)) + 
+  geom_vline(xintercept = p, color = 'red', linetype = 'dashed') +
+  geom_line() + 
+  labs(x = 'Probability',
+       y = 'Likelihood')
+
+```
+
+An inductive behavior approach was developed by J.
+Neyman and E.
+S.
+Pearson to address errors [@neyman1933].
+Neyman's approach emphasized that likelihoods were random and that further testing was needed so that inferences are not often wrong from true data generating functions.
+This development came to be the hypothesis test, where the level of error given certain assumptions about the data can be controlled.
+Many modern statistical methods incorporate some concept of the hypothesis test [@stroup] and has a central role in statistical inference.
 
 The Frequentist analysis technique begins with an underlying assumption of the data generating function.
-In practice, this function is rarely known, and thus is approximated by a reasonable probability function.
-Some examples include a Poisson distribution for count data, a binomial distribution for proportions of successes, and a normal distribution for unimodal continuous observations.
+In practice, this function is rarely known, and thus is approximated by a reasonable probability function or by attempting different models.
+For example, a generalized linear model can assume a Poisson or Negative Binomial distribution for data that has unrestricted counts, in contrast to the Binomial distribution where the total number of counts is known.
 
-After data is collected, a hypothesis test can be formulated.
-The general format of a Frequentist hypothesis test is to assume that a function of the data arose from an assumed data generating function.
-If the function of the data does not
+After data is collected, a hypothesis test can be formulated [@neyman1933].
+The idea of a Frequentist hypothesis test is to assume that a realization of the data arose from an assumed data generating function, called the null hypothesis.
+If the function of the data does not seem plausible under the null hypothesis, typically denoted as a p-value, then the null hypothesis is rejected in favor of the alternative hypothesis, which encompasses the complement of the null hypothesis.
 
-## Bayesian {#sec-bayesian}
+$$
+H_0:\theta=\theta_0 \\ H_1:\theta=\theta_1
+$$ {#eq-freq-ht}
 
-Bayesian inference begins with some knowledge about the parameter of interest.
+## Bayesian Inference {#sec-bayesian}
 
-## Visual {#sec-visual}
+Bayesian inference typically begins with some knowledge about the parameter of interest, although uninformative prior information can also be used.
+This knowledge is use used to formulate the posterior distribution, $\pi(\theta|X)$, which is the normalized product of the data likelihood under the parameter, $f(X|\theta)$, and the prior distribution, $\pi(\theta)$ (@eq-post).
+The posterior distribution is the main target of Bayesian inference, allowing the parameter of interest to exist as a probability function generated from past and current knowledge [@gelman2013].
 
-Unlike Frequentist and Bayesian inference, visual inference relies on the nuances of human perception.
-With visual inference, data is presented in the visual space and treated as a statistic.
-For example, a common assumption for linear models in the form of $Y=X\beta+\epsilon$ is that the residual term, $\epsilon$, is independently and identically distributed with respect to a Normal distribution with a mean of zero and an unknown variance.
-Formal tests, like the Shapiro-Wilk test \[, can be used to
+$$
+\pi(\theta|X) = \frac{f(X|\theta)\pi(\theta)}{\int f(X|\theta)\pi(\theta)d\theta}\propto f(X|\theta)\pi(\theta)
+$$ {#eq-post}
 
-The null data generation process
+Unlike Frequentist methodology, there are no p-values in Bayesian hypothesis testing, at least in the sense of the Frequentist definition for a p-value [@gelman2013; @held2018] .
+Instead, the Bayes factor can be used as evidence for competing hypotheses.
+The Bayes factor is the ratio of the posterior odds and prior odds, where $\theta_0$ denotes the parameter model proposed for one model, and $\theta_1$ is for an alternative model (@eq-bayes-factor).
+Bayes factors less than 0.1 are generally considered indicative of evidence against the model for $\theta_0$ [@held2018].
 
-One potential issue with visual inference is the human nature of wanting to find patterns in randomness.
+$$
+\text{Bayes Factor}=\frac{\pi(\theta_0|X)/\pi(\theta_1|X)}{\pi(\theta_0)/\pi(\theta_1)}
+$$ {#eq-bayes-factor}
+
+To illustrate a simple example of Bayesian inference, consider @fig-freq-binom where the true data generating function is $\text{Binomial}(n=20, p=0.7)$.
+The observed number of successes was 13 out of 20.
+Using an uninformative prior, the posterior distribution becomes @eq-binom-unif.
+
+$$
+\pi(p|X)) = {20 \choose x}p^x(1-p)^{n-x}\times 1\\ \propto p^{(x+1)-1}(1-p)^{(n-x+1)-1}\\ \propto Beta(x+1,n-x+1)
+$$ {#eq-binom-unif}
+
+Now suppose that it is known that previous data shows that $p$ is around 0.7.
+A sensible choice for the prior distribution is a Beta distribution, which will result in a closed-form solution.
+This is called a conjugate prior.
+
+$$
+\pi(p|X))  \propto p^{x}(1-p)^{n-x}\times p^{\alpha-1}(1-p)^{\beta-1}
+\\ =p^{(x+\alpha+1)-1}(1-p)^{(n-x+\beta+1)-1}
+\\ \propto Beta(x+\alpha+1,n-x+\beta+1)
+$$
+
+Letting $\alpha=9$ and $\beta=3.857$, the prior distribution has a mean of 0.7 and a variance of 0.01515. The posterior distribution becomes $\text{Beta}(x+10,n-x+4.857)$
+
+```{r}
+#| label: fig-bayes-binom
+#| fig-cap: "Caption"
+#| fig-subcap: 
+#| - "a: likelihoods"
+#| - "b: data simulated from models"
+#| layout-ncol: 2
+#| out-width: 90%
+
+set.seed(1999)
+n=20
+p=.7
+y=rbinom(1, n, p)
+
+alpha = 9
+beta = 3.857
+
+prior <- c('Uniform'='solid',
+           'Beta' = 'dashed')
+
+tibble(
+  x = seq(0,1, l = 1000)
+) %>% 
+  mutate(fx.u = dbeta(x, 13+1, 20-13+1),
+         fx.b = dbeta(x, 13+alpha+1, 20-13+beta+1),
+         L = choose(20,13)*x^(13)*(1-x)^(20-13)) %>% 
+  ggplot(mapping = aes(x = x)) + 
+  geom_line(aes(y = fx.u, linetype = 'Uniform')) + 
+  geom_line(aes(y = fx.b, linetype = 'Beta')) +
+  geom_vline(xintercept = 0.7, color = 'red') +
+  scale_linetype_manual(values = prior) + 
+  labs(x = 'p', y = expression(pi))
+
+```
+
+## Visual Inference {#sec-visual}
+
+In contrast to Bayesian and Frequentist methodologies, visual inference is not a strictly mathematical formulation but rather the product of human perception.
+Visualizations are widely used in statistics, having proven useful in data exploration [@tukey1965; @tukeyEDA; @beniger1978] and checking model diagnostics [@loy2016].
+However, the use of visualizations for formal statistical inference is a relatively new development in the past two decades [@buja2009; @loy2015; @vanderplas2017].
+
+The primary method of applying visual inference is through the lineup protocol [@buja2009].
+In the lineup protocol, viewers are presented with a series of graphs that display either the target data or data generated according a null model.
+The viewers are then tasked with identifying the graph that is most different from the others.
+If viewers can identify the target graph, then there is evidence that the target graph is different than the graphs created under the null model.
+An example is provided in @fig-lineup-examp1.
+
+```{r}
+#| label: fig-lineup-examp1
+#| fig-height: 6
+#| fig-width: 6
+#| fig-cap: "One trial of a visual inference lineup study.
+In this series of graphs, one dataset was produced differently than the other datasets.
+Can you figure out which graph it is?
+The answer can be found in the discussion section of this paper."
+
+lineup1 <- function(n=100, m=20, answer = 8){
+  x = c(rnorm(n*(m)))
+  y = c(rgamma(n*(answer-1), 2, 1), runif(n), rgamma(n*(m-answer), 2, 1))
+  verify = c(rep('null', n*(answer-1)), rep('answer', n), rep('null', n*(m-answer)))
+  data.frame(x = x, y = y, set = rep(1:m, each = n), verify = verify)
+
+}
+
+set.seed(219031)
+ggplot(data = lineup1(), aes(x = x, y = y)) + 
+  geom_point(size = 1/2) + 
+  labs(x = '', y = '') +
+  facet_wrap(~set, scales = 'free') + 
+  theme(aspect.ratio = 1, axis.text = element_blank(),
+        axis.ticks = element_blank(),
+        panel.grid = element_blank())
+
+```
+
+Defining the target plot and null plots are essential steps in designing a lineup study.
+The target plot is chosen so that it reflects a specific condition.
+In contrast, the null plots are constructed to reflect conditions that the target graph do not satisfy.
+For example, linear models often have the assumption that residuals are independently and identically distributed with mean zero and a constant variance [@majumder2013].
+In this case, testing the residual assumption would involve using a linear model fit to the true data as the target plot.
+Data can be simulated according the true linear model, which would result in the null plots.
+If viewers can identify the residual plot of the true dataset, then visual inference would indicate that the independent and identically distributed assumption of the residuals is violated.
+See @fig-resid-examp for an example.
+
+```{r}
+#| label: fig-resid-examp
+#| layout-ncol: 2
+#| fig-cap: "A comparison of two residual plots for the model $y=X \\beta + \\epsilon$, where $\\epsilon \\sim iid N(0, \\sigma^2)$.
+In panel (a), data is generated so that the residual assumptions holds.
+Panel (b) violates the residual assumptions by including dependence though an increasing slope."
+#| fig-subcap: 
+#| - "Zero slope model"
+#| - "Non-zero slope model"
+#| out-width: 90%
+#| fig-width: 6
+
+set.seed(3021392)
+ggplot(mapping = aes(x = runif(100), y = scale(rnorm(100)))) + 
+  geom_point() + 
+  geom_smooth(method = 'lm') + 
+  theme(aspect.ratio = 1) + 
+  labs(x = 'Fitted', y = 'Residual')
+
+x = runif(100)
+y = 2*x + rnorm(100)
+ggplot(mapping = aes(x = x, y = scale(y))) + 
+  geom_point() + 
+  geom_smooth(method = 'lm') + 
+  theme(aspect.ratio = 1) + 
+  labs(x = 'Fitted', y = 'Residual')
+
+```
+
+The lineup protocol tests the following hypotheses.
+
+-   Null hypothesis: the target graph cannot be identified among graphs generated with a null model.
+
+-   Alternative hypothesis: the target graph can be identified among graphs generated with a null model.
+
+Assuming that a total of $m-1$ null plots, $1$ target plot, and $n$ participants, a Binomial distribution can be used to test the hypothesis that the target plot is chosen correctly:
+
+$$
+H_0: \pi=1/m \quad \text{vs.} \quad H_A: \pi\neq1/m
+$$
+
+However, this hypothesis test assumes that plots are chosen at random by viewers.
+In reality, viewers are searching for particular features in the plots to use as justification for their selection [@buja2009].
+A different method of evaluation is presented by @vanderplas2021, where plot selection dependencies can be modeled with a Dirichlet-multinomial distribution.
+
+# Comparison of Inferences
+
+Frequentist vs. Bayes
+
+-   Parameter fixed vs. random
+
+-   Types of statements
+
+-   Frequentist test statistics vs. posterior for Bayes
+
+-   p-values vs. Bayes factor
+
+-   Credible vs. confidence
+
+The differences between Frequentist and Bayesian inference mainly arise from how population parameters are defined [@pek; @acree2021].
+Frequentist inference assumes that a parameter for a well-defined population at a given time is a fixed and unknown value, where a function of the data is the random variable.
+In some cases, this is a reasonable assumption, such as estimating the average weight of smallmouth bass in Lake Pepin, Minnesota [@pepin].
+**Here, there is a true average weight for all smallmouth bass in the lake, but the value can only be estimated.** In contrast, Bayesian inference interprets the population parameter as a random variable to quantify uncertainty.
+For the case of smallmouth bass in Lake Pepin, prior information about average weights obtained from historical records can be combined with current data to formulate the posterior distribution.
+
+Treating the population parameter as fixed or random affects the types of statements that Frequentist and Bayesian inferences are allowed to make about the population.
+Under Frequentist methodology, functions of the data are considered random, and thus statements about the population parameter pertain to the sampling distribution of the data.
+In essence, this means that Frequentist statements answer the question "did my data capture the true population parameter?" For Bayesian methodology, the posterior distribution allows statements to be made about the population parameter itself.
+
+A useful tool for both Frequentist and Bayesian inferences is a range of plausible values for the population parameter [@pek].
+For Freqentists, this range of values is referred to as a confidence interval.
+The interpretations of confidence intervals are about how often repeated sampling of the population would result in the interval covering the true population parameter.
+These statements become "with X% confidence, the true population parameter is between *lower limit* and *upper limit*." The Bayesian use of the posterior distribution introduces the credible interval.
+With the credible interval, statements can be made directly about the population parameter, such as "there is X% probability that the population parameter is between *lower limit* and *upper limit.*"
+
+Due to the differences in how the population parameter is defined, the scope of inference drastically changes between Frequentist and Bayesian methods.
+Frequentists establish a hypothesis test based on a test statistic from the sampling distribution.
+Probabilities of the test statistic under the null hypothesis give evidence to either reject or fail to reject the null hypothesis.
+The scope of inference under Bayesian inference MORE HERE....
+
+```{r}
+#| label: fig-freq-bayes-ci
+#| fig-cap: "Caption"
+#| fig-subcap: 
+#| - "panel a"
+#| - "panel b"
+
+freq.ci <- data.frame()
+for(i in 1:100){
+  t.test.save <- t.test(rnorm(100))
+  tmp <- data.frame(mean = t.test.save$estimate,
+                    lcl = t.test.save$conf.int[1],
+                    ucl = t.test.save$conf.int[2],
+                    id = i)
+  freq.ci <- bind_rows(freq.ci, tmp)
+  rm(t.test.save)
+  rm(tmp)
+}
+freq.ci %>% 
+  mutate(contain0 = ifelse(0 > lcl & 0 <= ucl, 'yes', 'no')) %>% 
+  ggplot(mapping = aes(x = mean, y = id, color = contain0)) + 
+  geom_point() + 
+  geom_errorbar(aes(xmin = lcl, xmax = ucl)) + 
+  geom_vline(xintercept = 0, color = 'red', linetype = 'dashed')
+
+ggplot(data.frame(x = c(-6, 6)), aes(x = x)) +
+  stat_function(fun = dnorm, geom = 'area') +
+  theme_minimal() +
+  labs(title = "Plot of y = x^2", x = "x", y = "y")
+
+```
+
+Frequentist vs. Visual
+
+-   Visual as good as Frequentist
+
+-   Same type of hypothesis test
+
+-   Visual only has one realization in a single lineup
+
+The formulation of formal visual inference is rooted in the methodology of Frequentist inference [@buja2009; @hofmann2012; @vanderplas2021].
+Both methodologies assume a hypothesis test where evidence is provided through data as to reject or fail to reject the null hypothesis.
+Whereas Frequentist inference establishes evidence through a test statistic of the data, visual inference establishes evidence through the detection of human perception.
+This is an important distinction, since the null hypothesis of Frequentist testing using a probability function and visual tests use $m-1$ realizations of the null hypothesis.
+
+@majumder2013 showed that lineup studies can be comparable to traditional statistical tests.
+
+Bayes vs. Visual
+
+-   Visual good for checking data, see if generated data like actual data
+
+Bayesian and visual inferences are perhaps the most different from each other.
+Bayesians rely on prior information to construct a model for the population parameter, but visual inference is mostly concerned with detecting differences between null plots and the target plot.
+Of course, some testing procedures in visual inference rely on previous knowledge that not all null plots are created equally, and thus different signal strengths within the null plots can be incorporated in statistical significance calculations [@vanderplas2021].
+
+Another paragraph about visual tools for Bayes model checking
+
+While there are many differences between Frequentist and Bayesian testing, visual inference tends to be a useful tool for model diagnostic checks [@majumder2013; @loy2015; @loy2017].
+To oversimplify the field of statistics, the main objective is to produce models that are not so far off from the truth so that generalizations can be made about findings in nature.
+To quote George Box, "all models are wrong, but some are useful." This means that the process of creating null plots from a model and determining if the actual data can be distinguished is a useful tool to validate the fit of a statistical model.
+
+```{r}
+#This is to show good and bad bayes priors
+
+```
 
@@
 
-To effectively demonstrate the three inference types, consider the following example.
+In the National Hockey League (NHL), games played at home are filled with energy and excitement for the home team.
+This is called having home-ice advantage.
+Consider the Minnesota Wild during their 2024-2025 season.
+Suppose that we wish to know if the Wild are playing with home-ice advantage, meaning that they win more games at home than they do for away games.
+Data is shown in @tbl-wild.
+
+```{r}
+#| label: tbl-wild
+
+wild <- read_csv('wild23-25.csv', show_col_types = F) %>% 
+  mutate(pt_diff = abs(gf-ga))
+wild %>% 
+  group_by(season, result, location) %>% 
+  count() %>% 
+  pivot_wider(names_from = location, values_from = n) %>% 
+  mutate(season = paste0(season-1, '-', season)) %>% 
+  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild')
+```
+
+The first step of any analysis should be data exploration.
+The Minnesota Wild won 48.1 percent of their home games and 63.6 percent of their away games as of February 28th, 2025.
+Immediately noticeable is that the Wild seem to perform better when not playing on home-ice.
+We also have information about point differentials that may be helpful, which is the difference in scores at the end of the game.
+
+The Frequentist analysis begins with the assumption that games are independently and identically distributed with a Binomial distribution.
+However, this assumption is not reasonable since the team's performance can change throughout the season.
+For example, Ryan Hartman received a major penalty on the February 1st, 2025 game against the Ottawa Senators.
+The penalty was severe enough for Hartman to receive a 10-game suspension, and thus changing a part of the team's dynamic for those games.
+
+```{r}
+options(contrasts = c("contr.sum", "contr.poly"))
+
+mod.freq <- glm(result ~ location*pt_diff, 
+                data = dplyr::filter(wild, season == 2025),
+     family = 'binomial')
+
+car::Anova(mod.freq, type = 'III')
+em <- emmeans(mod.freq, ~location*pt_diff, type = 'response',
+              at = list(pt_diff = 1:6))
+ggplot(data.frame(em), aes(x = factor(pt_diff), y = prob, color = location)) + 
+  geom_point(position = position_dodge(width = 1/4)) +
+
+  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
+                position = position_dodge(width = 1/4),
+                width = 1/4) + 
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
+  theme_bw() + 
+  scale_y_continuous(limits = c(0,1))
+```
+
+The Frequentist generalized linear model shows that the point differential is the only significant term, which means that the Wild do not seem to have a winning advantage by playing on home-ice.
+The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are more likely to lose games for larger point differentials than for smaller point differentials.
+
+To evaluate the fit of the Frequentist GLM, one solution is to use a lineup study.
+Here, data can be simulated such that wins and losses are randomly chosen from a Binomial distribution with a probability of success using the coefficients obtained from the fitted model.
+@fig-freq-examp-lineup shows a lineup of ten datasets produced by simulating wins and losses.
+In this lineup, probabilities are plotted along the point differential for both away and home games.
+It is difficult to identify the true dataset, which means that the Frequentist GLM appears to fit well.
+
+```{r}
+wild.freq <- wild %>% 
+  filter(season == 2025) %>% 
+  mutate(prob = predict(mod.freq, type = 'response'),
+         rownum = row_number(),
+         sim_win = map(prob, \(p)(rbinom(n = 10, size = 1, prob = p)))) %>% 
+  unnest(sim_win) %>% 
+  group_by(rownum) %>% 
+  mutate(dataset = 1:n(), 
+         display = ifelse(dataset == 3, result, sim_win))
+
+```
+
+```{r}
+#| label: fig-freq-examp-lineup
+#| fig-cap: "Lineup study using simulated wins and losses for the Minnesota Wild. Nine of the plots show data simulated from the fitted GLM, and one plot uses the actual data."
+
+library(modelr)
+wild.freq %>% 
+  group_by(dataset) %>% 
+  nest() %>% 
+  mutate(glm = map(data, \(x)(glm(result ~ location*pt_diff,
+                                  data = x, family = 'binomial')))) %>% 
+  mutate(preds = map(glm, predict, type = 'response')) %>% 
+  mutate(newdata = map2(data, glm, add_predictions, type = 'response')) %>% 
+  unnest(newdata) %>% 
+  ggplot(mapping = aes(x = pt_diff, y = result, color = location)) + 
+  geom_point(position = position_jitter(height = 0.01, width = 0.1),
+             alpha = 1/2) +
+stat_smooth(method = "glm",
+            formula = y ~ x,
+            se = F,
+            method.args = list(family=binomial)) + xlab("Point differential") +
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
+  facet_wrap(~dataset, nrow = 2) + 
+  # scale_y_continuous(limits = c(0,1)) + 
+  theme_bw()
+```
+
+The next approach uses the `brms` package in R to fit a Bayesian GLM to the data using the Minnesota Wild win-loss record for the 2023-2024 season.
+Two models are compared, one with uninformative (flat) priors on each model coefficient, and another that uses Normal priors based on a Frequentist GLM fit for the 2023-2024 season.
+The standard errors obtained from the Frequentist GLM coefficients were small, and thus multiplied by 2 for use in the Bayes GLM to allow for more flexibility.
+
+```{r}
+#| label: fig-true-props
+#| fig-height: 4
+#| fig-width: 6
+#| fig-cap: ""
+#| fig-subcap: 
+#| - ""
+#| - ""
+wild %>% 
+  group_by(season, location, result, pt_diff) %>% 
+  count() %>% 
+  pivot_wider(names_from = result, names_prefix = 'result',
+              values_from = n, values_fill = 0) %>% 
+  mutate(propwin = result1 / (result0+result1)) %>% 
+  filter(!(result0 == 0 & result1 == 0)) %>% 
+  ggplot(mapping = aes(x = pt_diff, y = propwin, color = location)) + 
+  geom_line() + 
+  geom_point() + 
+  scale_y_continuous(limits = c(0,1)) + 
+  labs(x = 'Point differential',
+       y = 'Proportion of wins') +
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
+  facet_wrap(~season) + 
+  theme_bw() + 
+  theme(aspect.ratio = 1) 
+```
+
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+#| cache: true
+
+library(brms)
+
+mod.prior <- glm(result ~ location*pt_diff, data = filter(wild, season == 2024))
+# summary(mod.prior)
+
+bayes.unif <- brm(result ~ location*pt_diff,
+    data = filter(wild, season == 2025),
+    save_all_pars = TRUE,
+    family = bernoulli())
+
+# default_prior(result ~ location*pt_diff,
+#     data = filter(wild, season == 2025))
+
+bayes.priors <- brm(result ~ location*pt_diff,
+    data = filter(wild, season == 2025),
+    prior = c(
+      set_prior("normal(-0.025652,2*0.113112)", class = "b", coef = "location1"),
+      set_prior("normal(-0.007458,2*0.042187)", class = "b", coef = "pt_diff"),
+      set_prior("normal(0.006828,2*0.042187)", class = "b", coef = "location1:pt_diff"),
+      set_prior("normal(0.490724,2*0.113112)", class = 'Intercept')
+    ),
+    # sample_prior = 'only',
+    save_all_pars = TRUE,
+    family = bernoulli())
+```
+
+```{r}
+#| message: false
+
+set.seed(283718921)
+bayes_factor(bayes.unif, bayes.priors)
+```
+
+The Bayes GLM coeffiecients using uninformative priors are similar to the Frequentist approach (@tbl-coef-compare).
+This result should be expected since prior knowledge was not used to construct the posterior distributions.
+Additionally, the coefficients obtained using informative priors have are smaller than the uninformative priors model and the Frequentist model, and decrease further when decreasing the standard errors of the prior distributions.
+
+Comparing the two Bayesian models, the Bayes Factor is computed to be approximately 1.
+Although the coefficients of the two models differ, there is not much evidence to prefer the informative priors over the uninformative priors.
+One discrepancy between the two Bayesian models is that 95% credible interval for the uninformative model considers the point differential to be a significant term (-1.03, -0.11), whereas the informative model does not (-0.23, 0.08).
+In this case, it may be better to consider additional prior distributions or use model averaging to get a more robust conclusion [@hinne].
+
+```{r}
+#| label: tbl-coef-compare
+data.frame(
+  Frequentist = coef(mod.freq),
+  `Flat priors` = fixef(bayes.unif)[,1],
+  `Informative priors` = fixef(bayes.priors)[,1]
+) %>% 
+  rownames_to_column(var = 'Term') %>% 
+  knitr::kable(digits = 3, caption = 'Coefficients of Frequentist and Bayesian methods for GLMs fit to the Minnesota Wild data.')
+
+```
+
+```{r}
+#| label: fig-bayes-probs
+#| layout-ncol: 2
+#| fig-cap: "Caption"
+#| fig-subcap: 
+#| - "Panel a"
+#| - "Panel b"
+
+set.seed(42)
+em.unif <- emmeans(bayes.unif, ~location*pt_diff,
+                   at = list(pt_diff = 1:6), type = 'response')
+em.priors <- emmeans(bayes.priors, ~location*pt_diff,
+                   at = list(pt_diff = 1:6), type = 'response')
+
+ggplot(data.frame(em.unif), aes(x = factor(pt_diff), y = response, color = location)) + 
+  geom_point(position = position_dodge(width = 1/4)) +
+
+  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD),
+                position = position_dodge(width = 1/4),
+                width = 1/4) + 
+  labs(x = 'Point differential', y = 'Predicted Probability') + 
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
+  theme_bw() + theme(aspect.ratio = 1) +
+  scale_y_continuous(limits = c(0,1))
+
+ggplot(data.frame(em.priors), aes(x = factor(pt_diff), y = response, color = location)) + 
+  geom_point(position = position_dodge(width = 1/4)) +
+
+  geom_errorbar(aes(ymin = lower.HPD, ymax = upper.HPD),
+                position = position_dodge(width = 1/4),
+                width = 1/4) + 
+  labs(x = 'Point differential', y = 'Predicted Probability') + 
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
+  theme_bw() + theme(aspect.ratio = 1) +
+  scale_y_continuous(limits = c(0,1))
+```
+
+Similar to the Frequentist case, visual inference can be used to examine model diagnostics for the Bayesian models. This process involves simulating data from the model and checking to see if
 
@@
 
-# Conclusion {#sec-conclusion}
+There is no single way to approach statistical inference.
+The commonality of statistical methods is that uncertainty is quantified through a random variable, but then methods and interpretations start to diverge.
 
-@gelman2013
+Frequentists rely on data as it is provided. This is a reasonable situation when conducting novel experiments or collecting data from new sources.
+
+Viz inference answer: 8

### 2025-03-03 paper-vis-inference/paper.qmd
Paper pre-proofread

@@
 title: "Statistical Inference"
-subtitle: "Frequentist, Bayesian, and Visual oh my"
+subtitle: "A comparative review of Frequentist, Bayesian, and Visual methods"
 author: Tyler Wiederich
@@
 
-The quantification of uncertainty is one of the key components behind statistical methodology [@acree2021; @lehman2008book] .
+The quantification of uncertainty is one of the key components behind statistical methodology [@acree2021; @lehmann2008book].
 This is useful in cases where population parameters are unknown or unable to be reasonably measured.
@@
 For example, Q-Q plots are a common diagnostic tool for checking normality assumptions [@loy2016].
-While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist.
+While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist, as long as a reference model is obtainable for null data generation.
 
-This paper is organized as follows. The foundations and uses of Frequentist, Bayesian, and visual inferences are organized XXXXXX
+This paper is organized as follows.
+The foundations and uses of Frequentist, Bayesian, and visual inferences are explained, including how each method should be interpreted.
+Then each method is compared to the other methods.
+Finally, an example using a Binomial generalized linear model is fitted with Frequentist and Bayesian methods, and evaluated using visual inference.
 
@@
 In practice, this assumption can be addressed through randomization [@neyman1977] or checking model diagnostics [@loy2016].
-
-Frequentists treat population parameters as fixed and unknown values, where data are random variables drawn from
 
@@
 
-R.
-A.
 Fisher believed that probability was deductive in nature and that probabilistic events could be formulated mathematically to produce a maximum likelihood function [@fisher1935].
-For example, the binomial distribution is given by $P(X=x)={n\choose x}p^x(1-p)^{n-x}$, where $X$ is the random variable for the number of successes out of $n$ trials given a probability $p$ for $x$ successes.
+For example, the Binomial distribution is given by $P(X=x)={n\choose x}p^x(1-p)^{n-x}$, where $X$ is the random variable for the number of successes out of $n$ trials given a probability $p$ for $x$ successes.
 The probability mass function comes from multiplying the number of combinations that a series of successes can take to the probabilities of individual successes and individual failures, a formulation of combinatorics.
@@
 #| label: fig-freq-binom
-#| fig-cap: "Binomial example of R.A.
-Fisher's inductive inference.
+#| fig-cap: "Binomial example of Fisher's inductive inference.
 A single observation from a Binomial distribution is drawn and observed to have 13 successes out of 20 trials, where the true probability of success is 0.7.
-The likelihood function is optimized at $p=0.65$ and the red line in (b) shows the true probability."
+The likelihood function is optimized at $p=0.65$, and the red line in (b) shows the true probability of success in the population."
 #| fig-subcap: 
@@
   labs(x = 'Number of successes',
-       y = 'Probability')
+       y = 'Probability') + 
+  theme(aspect.ratio = 1)
 
@@
   labs(x = 'Probability',
-       y = 'Likelihood')
+       y = 'Likelihood') + 
+  theme(aspect.ratio = 1)
 
@@
 
-An inductive behavior approach was developed by J.
-Neyman and E.
+An inductive behavior approach was developed by Neyman and E.
 S.
@@
 This development came to be the hypothesis test, where the level of error given certain assumptions about the data can be controlled.
-Many modern statistical methods incorporate some concept of the hypothesis test [@stroup] and has a central role in statistical inference.
+These tests include Type 1 and Type 2 errors, which are measures of uncertainty regarding how likely the hypothesis tests arrive at an incorrect conclusion.
 
-The Frequentist analysis technique begins with an underlying assumption of the data generating function.
+Many modern statistical methods incorporate some concept of the hypothesis test [@stroup], which begin with an underlying assumption of the data generating function.
 In practice, this function is rarely known, and thus is approximated by a reasonable probability function or by attempting different models.
@@
 
-After data is collected, a hypothesis test can be formulated [@neyman1933].
-The idea of a Frequentist hypothesis test is to assume that a realization of the data arose from an assumed data generating function, called the null hypothesis.
-If the function of the data does not seem plausible under the null hypothesis, typically denoted as a p-value, then the null hypothesis is rejected in favor of the alternative hypothesis, which encompasses the complement of the null hypothesis.
+When data is collected, a hypothesis test can be formulated [@eq-freq-ht].
+The idea of a Frequentist hypothesis test is to assume that a realization of the data arose from an assumed data generating function, called the null hypothesis [@neyman1933].
+If the function of the data does not seem plausible under the null hypothesis, then the null hypothesis is rejected in favor of the alternative hypothesis, which encompasses the complement of the null hypothesis.
 
 $$
-H_0:\theta=\theta_0 \\ H_1:\theta=\theta_1
+H_0:\theta=\theta_0 \qquad H_1:\theta=\theta_1
 $$ {#eq-freq-ht}
+
+One method to evaluate a hypothesis test is through a test statistic.
+Test statistics are formed so that their sampling distribution are known under the null hypothesis.
+For example, if the null hypothesis assume the random sample is from a normal distribution with mean $\mu$ and variance $\sigma^2$, then the sampling distribution of the sample mean $\bar{X}=\frac{1}{N}\sum_{i=1}^{N}x_i$ can be rewritten as $Z=\frac{\bar{X}-\mu}{\sigma/\sqrt{N}}$, which follows a standard normal distribution.
+Given the data, a probability can be calculated for observing the test statistic under the null hypothesis, which is called a p-value (@fig-freq-eval-methods-1).
+If the p-value is small, then it was unlikely to see the observed data under the null hypothesis, giving evidence for the alternative hypothesis.
+However, a large p-value does not give evidence that the data was generated from the null hypothesis since this was an assumed condition.
+
+Confidence intervals are another method of evaluating hypothesis tests [@neyman1937].
+These intervals for two-sided tests are calculated using the form: $\text{Statistic } \pm c\times\text{Standard Error}$, where $c$ is defined to produce a $C$ percent confidence level.
+Since the data is random, the interpretation of Frequentist confidence intervals is that $C$ percent of confidence intervals calculated from repeated sampling of the population will contain the true parameter value $C$ percent of the time (@fig-freq-eval-methods-2).
+Evaluating hypotheses with confidence intervals simply involve checking if the parameter value set in the null hypothesis falls within the interval.
+If the value does not fall in the interval, then there is evidence to reject the null hypothesis.
+
+```{r}
+#| label: fig-freq-eval-methods
+#| fig-cap: "Frequentist evaluation methods using test statistics and confidence intervals under a null hypothesis that uses a standard normal distribution.
+Panel (a) shows the area under a standard normal for the Z-score of the observed sample mean.
+Panel (b) provides 95% confidence intervals for Z-scores of the sample mean across multiple samples."
+#| fig-subcap: 
+#| - "Probability of observing test statistic under null hypothesis."
+#| - "95% confidence interval for multiple repeated samples."
+#| out-width: 90%
+
+set.seed(20244)
+ggplot() + 
+  stat_function(fun = dnorm) + 
+  stat_function(fun = dnorm, xlim = c(1.5,4), 
+                geom = 'area',
+                fill = 'red', color = 'black') + 
+  labs(x = 'Z', y = expression(phi(Z))) + 
+  scale_x_continuous(limits = c(-3, 3)) + 
+  theme(aspect.ratio = 1)
+
+freq.ci <- data.frame()
+for(i in 1:100){
+  t.test.save <- t.test(rnorm(100))
+  tmp <- data.frame(mean = t.test.save$estimate,
+                    lcl = t.test.save$conf.int[1],
+                    ucl = t.test.save$conf.int[2],
+                    id = i)
+  freq.ci <- bind_rows(freq.ci, tmp)
+  rm(t.test.save)
+  rm(tmp)
+}
+freq.ci %>% 
+  mutate(contain0 = ifelse(0 > lcl & 0 <= ucl, 'yes', 'no')) %>% 
+  ggplot(mapping = aes(x = mean, y = id, color = contain0)) + 
+  geom_point() + 
+  geom_errorbar(aes(xmin = lcl, xmax = ucl)) + 
+  scale_color_brewer(palette = 'Set1') +
+  labs(x = 'Z', y = 'Sample ID', color = 'CI contain zero?') + 
+  geom_vline(xintercept = 0, color = 'red', linetype = 'dashed') + 
+  theme(aspect.ratio = 1)
+```
 
@@
 
-Bayesian inference typically begins with some knowledge about the parameter of interest, although uninformative prior information can also be used.
-This knowledge is use used to formulate the posterior distribution, $\pi(\theta|X)$, which is the normalized product of the data likelihood under the parameter, $f(X|\theta)$, and the prior distribution, $\pi(\theta)$ (@eq-post).
-The posterior distribution is the main target of Bayesian inference, allowing the parameter of interest to exist as a probability function generated from past and current knowledge [@gelman2013].
+Bayesian inference typically begins with some knowledge about the parameter of interest, although uninformative prior information can be used [@gelman2013; @held2018].
+This knowledge is used to formulate the posterior distribution, $\pi(\theta|X)$, which is the normalized product of the data likelihood under the parameter, $f(X|\theta)$, and the prior distribution, $\pi(\theta)$ (@eq-post).
+The posterior distribution is the main target of Bayesian inference, allowing the parameter of interest to exist as a probability function generated from past and current knowledge.
 
@@
 The observed number of successes was 13 out of 20.
-Using an uninformative prior, the posterior distribution becomes @eq-binom-unif.
+Using an uninformative prior, the posterior distribution becomes @eq-binom-unif, which is proportional to a Beta distribution with $\alpha=x+1$ and $\beta=n-x+1$.
 
 $$
-\pi(p|X)) = {20 \choose x}p^x(1-p)^{n-x}\times 1\\ \propto p^{(x+1)-1}(1-p)^{(n-x+1)-1}\\ \propto Beta(x+1,n-x+1)
+\pi(p|X)) = {20 \choose x}p^x(1-p)^{n-x}\times 1 \propto p^{(x+1)-1}(1-p)^{(n-x+1)-1}
 $$ {#eq-binom-unif}
@@
 A sensible choice for the prior distribution is a Beta distribution, which will result in a closed-form solution.
-This is called a conjugate prior.
+This is called a conjugate prior [@raiffa2000].
 
 $$
-\pi(p|X))  \propto p^{x}(1-p)^{n-x}\times p^{\alpha-1}(1-p)^{\beta-1}
-\\ =p^{(x+\alpha+1)-1}(1-p)^{(n-x+\beta+1)-1}
-\\ \propto Beta(x+\alpha+1,n-x+\beta+1)
+\pi(p|X))  \propto p^{x}(1-p)^{n-x}\times p^{\alpha-1}(1-p)^{\beta-1} \propto Beta(x+\alpha+1,n-x+\beta+1)
 $$
 
-Letting $\alpha=9$ and $\beta=3.857$, the prior distribution has a mean of 0.7 and a variance of 0.01515. The posterior distribution becomes $\text{Beta}(x+10,n-x+4.857)$
+Letting $\alpha=9$ and $\beta=3.857$, the prior distribution has a mean of 0.7 and a variance of 0.01515.
+The posterior distribution becomes $\text{Beta}(x+10,n-x+4.857)$, which shifts the posterior closer to 0.7 than when using the uniform prior (@fig-bayes-binom).
+
+Similar to Frequentist methods, intervals of the parameters can be created for the posterior distribution, called credible intervals [@acree2021].
+These intervals are calculated simply by encompassing a desired probability from the posterior distribution.
+The result is that interpretations are directly about the parameter and not repeated sampling of the data.
+From the binomial example with a uniform prior, the two-sided 95% credible interval states that there is a 95% probability that the proportion of success is between 0.43 and 0.82.
 
@@
   scale_linetype_manual(values = prior) + 
-  labs(x = 'p', y = expression(pi))
+  labs(x = 'p', y = 'Posterior density', linetype = 'Prior')
 
@@
 
+Selection of an appropriate prior distribution is an important part of Bayesian methods [@pek2020; @gelman2013; @strachan2003].
+Prior distributions are typically chosen so that the strength of belief, or lack thereof, is accounted for.
+Overly optimistic priors either favor certain results or have smaller variances around anticipated values.
+When information is not available or there is a weak belief in prior information, a uniform prior or a prior with a large variance can be used.
+However, the choice of models can be evaluated against each other [@held2018] or averaged together [@hoeting1999; @hinne2020]
+
 ## Visual Inference {#sec-visual}
 
-In contrast to Bayesian and Frequentist methodologies, visual inference is not a strictly mathematical formulation but rather the product of human perception.
 Visualizations are widely used in statistics, having proven useful in data exploration [@tukey1965; @tukeyEDA; @beniger1978] and checking model diagnostics [@loy2016].
 However, the use of visualizations for formal statistical inference is a relatively new development in the past two decades [@buja2009; @loy2015; @vanderplas2017].
+In contrast to Bayesian and Frequentist methodologies, visual inference is not a strictly mathematical formulation but rather the product of human perception.
 
 The primary method of applying visual inference is through the lineup protocol [@buja2009].
-In the lineup protocol, viewers are presented with a series of graphs that display either the target data or data generated according a null model.
+In the lineup protocol, viewers are presented with a series of graphs that display the target data and data generated according a null model.
 The viewers are then tasked with identifying the graph that is most different from the others.
@@
 In this case, testing the residual assumption would involve using a linear model fit to the true data as the target plot.
-Data can be simulated according the true linear model, which would result in the null plots.
+Data can be simulated according the estimated linear model and fitted to the same model, which would result in the null plots.
 If viewers can identify the residual plot of the true dataset, then visual inference would indicate that the independent and identically distributed assumption of the residuals is violated.
@@
 In panel (a), data is generated so that the residual assumptions holds.
-Panel (b) violates the residual assumptions by including dependence though an increasing slope."
+Panel (b) violates the residual assumptions by including non-constant variance."
 #| fig-subcap: 
-#| - "Zero slope model"
-#| - "Non-zero slope model"
+#| - "Independent residuals"
+#| - "Dependent residuals"
 #| out-width: 90%
@@
 x = runif(100)
-y = 2*x + rnorm(100)
+y = rnorm(100, sd = x)
 ggplot(mapping = aes(x = x, y = scale(y))) + 
@@
 
-The lineup protocol tests the following hypotheses.
-
--   Null hypothesis: the target graph cannot be identified among graphs generated with a null model.
-
--   Alternative hypothesis: the target graph can be identified among graphs generated with a null model.
-
+The lineup protocol tests the null hypothesis that the target graph cannot be identified among graphs generated with a null model against the alternative hypothesis that the target graph can be identified among graphs generated with a null model.
 Assuming that a total of $m-1$ null plots, $1$ target plot, and $n$ participants, a Binomial distribution can be used to test the hypothesis that the target plot is chosen correctly:
@@
 
-Frequentist vs. Bayes
-
--   Parameter fixed vs. random
-
--   Types of statements
-
--   Frequentist test statistics vs. posterior for Bayes
-
--   p-values vs. Bayes factor
-
--   Credible vs. confidence
+## Frequentist vs. Bayesian
 
 The differences between Frequentist and Bayesian inference mainly arise from how population parameters are defined [@pek; @acree2021].
-Frequentist inference assumes that a parameter for a well-defined population at a given time is a fixed and unknown value, where a function of the data is the random variable.
-In some cases, this is a reasonable assumption, such as estimating the average weight of smallmouth bass in Lake Pepin, Minnesota [@pepin].
-**Here, there is a true average weight for all smallmouth bass in the lake, but the value can only be estimated.** In contrast, Bayesian inference interprets the population parameter as a random variable to quantify uncertainty.
-For the case of smallmouth bass in Lake Pepin, prior information about average weights obtained from historical records can be combined with current data to formulate the posterior distribution.
+Frequentist inference assumes that a parameter for a well-defined population at a given time is fixed and unknown, where a function of the data is the random variable.
+In some cases, this is a reasonable assumption, such as estimating the average weight of smallmouth bass in Lake Pepin, Minnesota \[[www.dnr.state.mn.us/lakefind/](https://www.dnr.state.mn.us/lakefind/showreport.html?downum=25000100)\](<https://www.dnr.state.mn.us/lakefind/showreport.html?downum=25000100>).
+Here, there is a true average weight for all smallmouth bass in the lake, but the value can only be estimated.
+In contrast, Bayesian inference interprets the population parameter as a random variable to quantify uncertainty.
+For the case of smallmouth bass in Lake Pepin, prior information about average weights obtained from historical records can be combined with current data to formulate the posterior distribution to update beliefs on average weights.
 
@@
 Under Frequentist methodology, functions of the data are considered random, and thus statements about the population parameter pertain to the sampling distribution of the data.
-In essence, this means that Frequentist statements answer the question "did my data capture the true population parameter?" For Bayesian methodology, the posterior distribution allows statements to be made about the population parameter itself.
+In essence, this means that Frequentist statements answer the question "how often would this outcome occur under repeated sampling?" For Bayesian methodology, the posterior distribution allows statements to be made directly about the population parameters.
 
 A useful tool for both Frequentist and Bayesian inferences is a range of plausible values for the population parameter [@pek].
-For Freqentists, this range of values is referred to as a confidence interval.
+For Freqentists, this range of values is referred to as a confidence interval [@neyman1937].
 The interpretations of confidence intervals are about how often repeated sampling of the population would result in the interval covering the true population parameter.
 These statements become "with X% confidence, the true population parameter is between *lower limit* and *upper limit*." The Bayesian use of the posterior distribution introduces the credible interval.
-With the credible interval, statements can be made directly about the population parameter, such as "there is X% probability that the population parameter is between *lower limit* and *upper limit.*"
-
-Due to the differences in how the population parameter is defined, the scope of inference drastically changes between Frequentist and Bayesian methods.
-Frequentists establish a hypothesis test based on a test statistic from the sampling distribution.
-Probabilities of the test statistic under the null hypothesis give evidence to either reject or fail to reject the null hypothesis.
-The scope of inference under Bayesian inference MORE HERE....
+With the credible interval, statements can be made directly about the population parameter, such as "there is X% probability that the population parameter is between *lower limit* and *upper limit.*" @fig-freq-bayes-ci shows an example of how the intervals differ between Frequentists and Bayesians.
 
@@
 #| label: fig-freq-bayes-ci
-#| fig-cap: "Caption"
+#| fig-cap: "Comparison of interprations between confidence and credible intervals.
+Confidence intervals rely on repeated sampling of the data, whereas credible intervals use probabilities of the posterior distribution."
+#| out-width: 90%
 #| fig-subcap: 
-#| - "panel a"
-#| - "panel b"
+#| - "Confidence interval"
+#| - "Credible interval"
 
@@
   geom_errorbar(aes(xmin = lcl, xmax = ucl)) + 
-  geom_vline(xintercept = 0, color = 'red', linetype = 'dashed')
+  scale_color_brewer(palette = 'Set1') +
+  labs(x = expression(theta), y = 'Sample ID', color = 'CI contain zero?') + 
+  geom_vline(xintercept = 0, color = 'red', linetype = 'dashed') + 
+  theme(aspect.ratio = 1)
 
-ggplot(data.frame(x = c(-6, 6)), aes(x = x)) +
-  stat_function(fun = dnorm, geom = 'area') +
+ggplot(data.frame(x = c(-4, 4)), aes(x = x)) +
+  stat_function(fun = dnorm, geom = 'line') +
+  stat_function(fun = dnorm, geom = 'area', xlim = c(-2, 2), fill = '#377EB8FF', color = 'black') +
   theme_minimal() +
-  labs(title = "Plot of y = x^2", x = "x", y = "y")
+  labs(x = expression(theta), y = 'Posterior density') + 
+  theme(aspect.ratio = 1)
 
@@
 
-Frequentist vs. Visual
+Due to the differences in how the population parameter is defined, the scope of inference drastically changes between Frequentist and Bayesian methods.
+Frequentists establish a hypothesis test based on a test statistic from the sampling distribution.
+Probabilities of the test statistic under the null hypothesis give evidence to either reject or fail to reject the null hypothesis.
+For Bayesian methods, hypothesis testing is redundant since the posterior contains all of the necessary information about the population parameter, including the estimate and credible interval.
+The focus of Bayesian inference is about obtaining well-fitted models from reasonably chosen prior distributions.
 
--   Visual as good as Frequentist
-
--   Same type of hypothesis test
-
--   Visual only has one realization in a single lineup
+## Frequentist vs. Visual
 
@@
 This is an important distinction, since the null hypothesis of Frequentist testing using a probability function and visual tests use $m-1$ realizations of the null hypothesis.
+However, @majumder2013 showed that lineup studies can be comparable to traditional statistical tests.
 
-@majumder2013 showed that lineup studies can be comparable to traditional statistical tests.
-
-Bayes vs. Visual
-
--   Visual good for checking data, see if generated data like actual data
+## Bayesian vs. Visual
 
@@
 Of course, some testing procedures in visual inference rely on previous knowledge that not all null plots are created equally, and thus different signal strengths within the null plots can be incorporated in statistical significance calculations [@vanderplas2021].
+This approach is not directly a Bayesian inference, but can lead into a Bayesian approach by using the relationship between the multinomial and Dirichlet distributions.
 
-Another paragraph about visual tools for Bayes model checking
+## Visual Inference as a Diagnostic Tool
 
@@
 To oversimplify the field of statistics, the main objective is to produce models that are not so far off from the truth so that generalizations can be made about findings in nature.
-To quote George Box, "all models are wrong, but some are useful." This means that the process of creating null plots from a model and determining if the actual data can be distinguished is a useful tool to validate the fit of a statistical model.
+This means that the process of creating null plots from a model and determining if the actual data can be distinguished is a useful tool to validate the fit of a statistical model.
 
-```{r}
-#This is to show good and bad bayes priors
+The use of visual inference diagnostic checks for Frequentist methods involve generating data from a fitted model.
+After data is generated, one possible diagnostic check is to place the true data in lineup study with the generated data.
+If the true data is indistinguishable from the simulated data, then it is reasonable to assume that the model fits well.
+Another option is to use the residual plots for models refitted to the simulated data compare them to the actual residuals.
+The residual plots can be useful when the usual residual assumption of normality with mean 0 and constant variance is not modeled, such as for repeated measures experiments.
 
-```
+For Bayesian analyses, the process of incorporating visual inference is slightly different than for Frequentist analyses.
+First, parameter values need to be drawn from the posterior distribution(s).
+Then data can be simulated and used in a lineup study with the observed data as the target plot.
+This process could also be extended to comparing two posterior distributions that use different priors, similar to how the Bayes factor measures evidence for favoring one model over another.
 
@@
 Consider the Minnesota Wild during their 2024-2025 season.
-Suppose that we wish to know if the Wild are playing with home-ice advantage, meaning that they win more games at home than they do for away games.
+Suppose that the head coach wishes to know if the Wild are playing with an advantage playing on home-ice, meaning that they win more games at home than they do for away games.
 Data is shown in @tbl-wild.
@@
   mutate(season = paste0(season-1, '-', season)) %>% 
-  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild')
+  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. 0 represents losses and 1 represents wins.')
 ```
 
-The first step of any analysis should be data exploration.
+The first step of any analysis should be data exploration (@fig-true-props).
 The Minnesota Wild won 48.1 percent of their home games and 63.6 percent of their away games as of February 28th, 2025.
@@
 We also have information about point differentials that may be helpful, which is the difference in scores at the end of the game.
-
-The Frequentist analysis begins with the assumption that games are independently and identically distributed with a Binomial distribution.
-However, this assumption is not reasonable since the team's performance can change throughout the season.
-For example, Ryan Hartman received a major penalty on the February 1st, 2025 game against the Ottawa Senators.
-The penalty was severe enough for Hartman to receive a 10-game suspension, and thus changing a part of the team's dynamic for those games.
-
-```{r}
-options(contrasts = c("contr.sum", "contr.poly"))
-
-mod.freq <- glm(result ~ location*pt_diff, 
-                data = dplyr::filter(wild, season == 2025),
-     family = 'binomial')
-
-car::Anova(mod.freq, type = 'III')
-em <- emmeans(mod.freq, ~location*pt_diff, type = 'response',
-              at = list(pt_diff = 1:6))
-ggplot(data.frame(em), aes(x = factor(pt_diff), y = prob, color = location)) + 
-  geom_point(position = position_dodge(width = 1/4)) +
-
-  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
-                position = position_dodge(width = 1/4),
-                width = 1/4) + 
-  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
-  theme_bw() + 
-  scale_y_continuous(limits = c(0,1))
-```
-
-The Frequentist generalized linear model shows that the point differential is the only significant term, which means that the Wild do not seem to have a winning advantage by playing on home-ice.
-The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are more likely to lose games for larger point differentials than for smaller point differentials.
-
-To evaluate the fit of the Frequentist GLM, one solution is to use a lineup study.
-Here, data can be simulated such that wins and losses are randomly chosen from a Binomial distribution with a probability of success using the coefficients obtained from the fitted model.
-@fig-freq-examp-lineup shows a lineup of ten datasets produced by simulating wins and losses.
-In this lineup, probabilities are plotted along the point differential for both away and home games.
-It is difficult to identify the true dataset, which means that the Frequentist GLM appears to fit well.
-
-```{r}
-wild.freq <- wild %>% 
-  filter(season == 2025) %>% 
-  mutate(prob = predict(mod.freq, type = 'response'),
-         rownum = row_number(),
-         sim_win = map(prob, \(p)(rbinom(n = 10, size = 1, prob = p)))) %>% 
-  unnest(sim_win) %>% 
-  group_by(rownum) %>% 
-  mutate(dataset = 1:n(), 
-         display = ifelse(dataset == 3, result, sim_win))
-
-```
-
-```{r}
-#| label: fig-freq-examp-lineup
-#| fig-cap: "Lineup study using simulated wins and losses for the Minnesota Wild. Nine of the plots show data simulated from the fitted GLM, and one plot uses the actual data."
-
-library(modelr)
-wild.freq %>% 
-  group_by(dataset) %>% 
-  nest() %>% 
-  mutate(glm = map(data, \(x)(glm(result ~ location*pt_diff,
-                                  data = x, family = 'binomial')))) %>% 
-  mutate(preds = map(glm, predict, type = 'response')) %>% 
-  mutate(newdata = map2(data, glm, add_predictions, type = 'response')) %>% 
-  unnest(newdata) %>% 
-  ggplot(mapping = aes(x = pt_diff, y = result, color = location)) + 
-  geom_point(position = position_jitter(height = 0.01, width = 0.1),
-             alpha = 1/2) +
-stat_smooth(method = "glm",
-            formula = y ~ x,
-            se = F,
-            method.args = list(family=binomial)) + xlab("Point differential") +
-  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
-  facet_wrap(~dataset, nrow = 2) + 
-  # scale_y_continuous(limits = c(0,1)) + 
-  theme_bw()
-```
-
-The next approach uses the `brms` package in R to fit a Bayesian GLM to the data using the Minnesota Wild win-loss record for the 2023-2024 season.
-Two models are compared, one with uninformative (flat) priors on each model coefficient, and another that uses Normal priors based on a Frequentist GLM fit for the 2023-2024 season.
-The standard errors obtained from the Frequentist GLM coefficients were small, and thus multiplied by 2 for use in the Bayes GLM to allow for more flexibility.
 
@@
 #| fig-cap: ""
-#| fig-subcap: 
-#| - ""
-#| - ""
+
 wild %>% 
@@
 
+The Frequentist analysis begins with the assumption that games are independently and identically distributed with a Binomial distribution.
+However, this assumption is not reasonable since the team's performance can change throughout the season.
+For example, Ryan Hartman received a major penalty on the February 1st, 2025 game against the Ottawa Senators.
+The penalty was severe enough for Hartman to receive a 10-game suspension and thus changing a part of the team's dynamic for those games.
+
+```{r}
+#| label: tbl-anova
+
+options(contrasts = c("contr.sum", "contr.poly"))
+
+mod.freq <- glm(result ~ location + I(pt_diff) + I(pt_diff^2), 
+                data = dplyr::filter(wild, season == 2025),
+     family = 'binomial')
+pander::pander(anova(mod.freq))
+
+```
+
+```{r}
+em <- emmeans(mod.freq, ~location*pt_diff, type = 'response',
+              at = list(pt_diff = 1:6))
+ggplot(data.frame(em), aes(x = factor(pt_diff), y = prob, color = location)) + 
+  geom_point(position = position_dodge(width = 1/4)) +
+
+  geom_errorbar(aes(ymin = asymp.LCL, ymax = asymp.UCL),
+                position = position_dodge(width = 1/4),
+                width = 1/4) + 
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
+  theme_bw() + 
+  scale_y_continuous(limits = c(0,1)) + 
+  labs(x = 'Point differential', y = 'Estimated probability of winning')
+```
+
+Fitting a Frequentist generalized linear model (GLM) as shown in @eq-freq-glm reveals that the point differential is the only significant term (@tbl-anova), which means that the Wild do not seem to have a winning advantage by playing on home-ice.
+The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are less likely to win games for larger point differentials than for smaller point differentials.
+
+$$
+\eta_{ij}=\log(\frac{\pi_{ij}}{(1-\pi_{ij})})=\beta_0+\beta_{1i}+\beta_2X_{j}+\beta_3X_{j}^2
+$$ {#eq-freq-glm}
+
+To evaluate the fit of the Frequentist GLM, one solution is to use a lineup study.
+Here, data can be simulated such that wins and losses are randomly chosen from a Binomial distribution with a probability of success using the coefficients obtained from the fitted model.
+@fig-freq-examp-lineup shows a lineup of ten datasets produced by simulating wins and losses.
+In this lineup, probabilities are plotted along the point differential for both away and home games.
+It is difficult to identify the true dataset, which means that the Frequentist GLM appears to fit well.
+
+```{r}
+set.seed(2025)
+wild.freq <- wild %>% 
+  filter(season == 2025) %>% 
+  mutate(prob = predict(mod.freq, type = 'response'),
+         rownum = row_number(),
+         sim_win = map(prob, \(p)(rbinom(n = 10, size = 1, prob = p)))) %>% 
+  unnest(sim_win) %>% 
+  group_by(rownum) %>% 
+  mutate(dataset = 1:n(), 
+         display = ifelse(dataset == 3, result, sim_win))
+
+```
+
+```{r}
+#| label: fig-freq-examp-lineup
+#| fig-cap: "Lineup study using simulated wins and losses for the Minnesota Wild.
+Nine of the plots show data simulated from the fitted GLM, and one plot uses the actual data.
+The target plot is plot number (81 divided by 27)."
+
+library(modelr)
+ wild.freq %>% 
+  group_by(dataset) %>% 
+  nest() %>% 
+  mutate(glm = map(data, \(x)(glm(display ~ location + pt_diff + I(pt_diff^2),
+                                  data = x, family = 'binomial')))) %>% 
+  mutate(preds = map(glm, predict, type = 'response')) %>% 
+  mutate(newdata = map2(data, glm, add_predictions, type = 'response')) %>% 
+  unnest(newdata) %>% 
+  ggplot(mapping = aes(x = pt_diff, y = result, color = location)) + 
+  geom_point(position = position_jitter(height = 0.01, width = 0.1),
+             alpha = 1/2) +
+   geom_line(aes(y = pred)) + 
+# stat_smooth(method = "glm",
+#             formula = y ~ poly(x,2, raw =T),
+#             se = F,
+#             method.args = list(family=binomial)) + xlab("Point differential") +
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
+  facet_wrap(~dataset, nrow = 2) + 
+  # scale_y_continuous(limits = c(0,1)) + 
+  theme_bw()
+```
+
+The Bayesian approach uses the `brms` package in R to fit a Bayesian GLM to the data using the Minnesota Wild win-loss record for the 2023-2024 season.
+Two models are compared, one with uninformative (flat) priors on each model coefficient, and another that uses normal priors based on a Frequentist GLM fit for the 2023-2024 season.
+The standard errors obtained from the Frequentist GLM coefficients were small, and thus multiplied by 2 for use in the Bayes GLM with normal priors to allow for more flexibility.
+
 ```{r}
@@
 
+set.seed(2013910)
 library(brms)
 
-mod.prior <- glm(result ~ location*pt_diff, data = filter(wild, season == 2024))
-# summary(mod.prior)
+mod.prior <- glm(result ~ location + pt_diff + I(pt_diff^2), data = filter(wild, season == 2024))
 
-bayes.unif <- brm(result ~ location*pt_diff,
+summary(mod.prior)
+
+bayes.unif <- brm(result ~ location + pt_diff + I(pt_diff^2),
     data = filter(wild, season == 2025),
@@
 
-# default_prior(result ~ location*pt_diff,
-#     data = filter(wild, season == 2025))
+default_prior(result ~ location + pt_diff + I(pt_diff^2),
+    data = filter(wild, season == 2025))
 
-bayes.priors <- brm(result ~ location*pt_diff,
+bayes.priors <- brm(result ~  location + pt_diff + I(pt_diff^2),
     data = filter(wild, season == 2025),
     prior = c(
-      set_prior("normal(-0.025652,2*0.113112)", class = "b", coef = "location1"),
-      set_prior("normal(-0.007458,2*0.042187)", class = "b", coef = "pt_diff"),
-      set_prior("normal(0.006828,2*0.042187)", class = "b", coef = "location1:pt_diff"),
-      set_prior("normal(0.490724,2*0.113112)", class = 'Intercept')
+      set_prior("normal(-0.4379,0.2310)", class = "b", coef = "Ipt_diffE2"),
+      set_prior("normal(0.3480,0.3002)", class = "b", coef = "location1"),
+      set_prior("normal(1.8803,1.1902)", class = "b", coef = "pt_diff"),
+      set_prior("normal(-1.0285,1.2716)", class = 'Intercept')
     ),
@@
 #| message: false
+#| include: false
 
 set.seed(283718921)
-bayes_factor(bayes.unif, bayes.priors)
+bayes_factor(bayes.priors, bayes.unif)
 ```
 
-The Bayes GLM coeffiecients using uninformative priors are similar to the Frequentist approach (@tbl-coef-compare).
-This result should be expected since prior knowledge was not used to construct the posterior distributions.
-Additionally, the coefficients obtained using informative priors have are smaller than the uninformative priors model and the Frequentist model, and decrease further when decreasing the standard errors of the prior distributions.
+The Bayes GLM coeffiecients for both models are similar to the Frequentist approach (@tbl-coef-compare).
+This result is expected since the uninformative model puts more weight on the data and the informative model has relatively small standard errors for its prior distributions.
+Additionally, the coefficients obtained using informative priors have have smaller standard errors than those obtained for the uninformative priors model and the Frequentist GLM.
 
-Comparing the two Bayesian models, the Bayes Factor is computed to be approximately 1.
-Although the coefficients of the two models differ, there is not much evidence to prefer the informative priors over the uninformative priors.
-One discrepancy between the two Bayesian models is that 95% credible interval for the uninformative model considers the point differential to be a significant term (-1.03, -0.11), whereas the informative model does not (-0.23, 0.08).
-In this case, it may be better to consider additional prior distributions or use model averaging to get a more robust conclusion [@hinne].
+Comparing the two Bayesian models and using the model with informative priors as the "null hypothesis", the Bayes Factor is computed to be approximately 0.42.
+Although the standard errors of the parameters for the two models differ, there is not much evidence to prefer the informative priors over the uninformative priors.
+One discrepancy between the two Bayesian models is that 95% credible interval for the uninformative model does not consider the linear point differential to be a significant term (-0.02, -4.84), whereas the informative model does (0.62, 3.33).
+In this case, it may be better to consider additional prior distributions or use model averaging to obtain a more robust conclusion [@hinne].
 
@@
 #| layout-ncol: 2
-#| fig-cap: "Caption"
+#| fig-cap: "Predicted probabilites under two Bayesian models. The model using normal priors has smaller credible intervals than the model using flat priors."
 #| fig-subcap: 
-#| - "Panel a"
-#| - "Panel b"
+#| - "Flat priors"
+#| - "Normal priors"
+#| out-width: 90%
 
@@
 
-Similar to the Frequentist case, visual inference can be used to examine model diagnostics for the Bayesian models. This process involves simulating data from the model and checking to see if
+Similar to the Frequentist case, visual inference can be used to examine model diagnostics for the Bayesian models.
+Here, it would be worth using comparing the uninformative and informative models to provide additional evidence along with the Bayes factor for which model to use since the linear term of point differential has conflicting conclusions.
 
@@
 There is no single way to approach statistical inference.
-The commonality of statistical methods is that uncertainty is quantified through a random variable, but then methods and interpretations start to diverge.
+The commonality of statistical methods is that uncertainty is quantified through a random variable, but defining the random variable is where the methods diverge in computation and interpretations.
+In this paper, Frequentist and Bayesian inferences were compared along with visual inference, and when each methodology should be used.
 
-Frequentists rely on data as it is provided. This is a reasonable situation when conducting novel experiments or collecting data from new sources.
+Frequentists rely on data as it is provided.
+This is a reasonable situation when conducting novel experiments or collecting data from new sources.
+Statements of uncertainty pertain to what happens under repeated sampling conditions, which is useful when wanting to know what conditions to expect for the next instance of data collection.
+Results are more easily calculated, having test statistics and confidence intervals based on sampling distributions of statistics from the parameter.
 
-Viz inference answer: 8
+In contrast to Frequentists, Bayesian inference is about updating beliefs given prior information.
+The behavior of the population parameter is treated as random and updated through a new instances of data collection.
+The goal is to estimate the posterior distribution so that statements can be made about the parameter(s) of interest.
+Other than model checking, this is where Bayesian analyses typically end since the goal is to make make inferences on the parameter(s).
+Bayesian methods are typically computationally demanding due to the complex posterior calculations.
+
+Lastly, visual inference is unique in that it can function on its own or in addition to Frequentist and Bayesian methods [@loy2017; @gelman2013].
+The main goal of lineup studies is to determine if a target plot can be identified.
+If it can be distinguished, then there is evidence that humans a particular feature of the target plot.
+From @fig-lineup-examp1, null plots were modeled with a gamma distribution for the y-axis variable and a normal distribution for the x-axis variable.
+The target plot was plot number 8, where the y-axis variable used a uniform distribution instead of a gamma distribution.
+Additionally, these lineup tests can be adapted for model fitting in Frequentist and Bayesian analyses, which can help to verify that models are reasonable for the collected data.
+
+Frequentist, Bayesian, and visual inference each have their own place in estimating phenomena of the natural world.
+After data collection, the type of analysis mostly comes down to personal preference and adequately fitting reasonable models.
+Frequentists use data as the foundation their analyses and Bayesians use data and prior beliefs.
+Visual inference exists as a method that can be used in combination with analyses, but also for inferences regarding data visualization.
+Picking the right type of inference is useful for providing data-driven results.
+
+### References
+
+::: {#refs}
+:::

### 2025-03-04 paper-vis-inference/paper.qmd
Changes before proofread

@@
 date: "`r format(Sys.time(), '%B %d, %Y')`"
-format: 
-  wordcount-pdf: default
+format: wordcount-pdf
 bibliography: references.bib
@@
   warning: false
+  include: false
 ---
@@
 
-The quantification of uncertainty is one of the key components behind statistical methodology [@acree2021; @lehmann2008book].
+The quantification of uncertainty is one of the fundamental concepts behind statistical methodology [@acree2021; @lehmann2008book].
 This is useful in cases where population parameters are unknown or unable to be reasonably measured.
-For example, some possible research questions that would require statistical methods include “What percentage of Minnesotan residents agree with the proposed legislation?,” or “Does the placement of higher margin products on the middle shelf of a store aisle lead to more profits than placement on top and bottom shelves?” These types of questions can be approximated using statistical inference, although the approach differs by the specific method of inference.
+For example, some possible research questions that would require statistical methods include “What percentage of Minnesotan residents agree with the proposed legislation?,” or “Does the placement of higher margin products on the middle shelf of a store aisle lead to more profits than placement on top and bottom shelves?” These types of questions can be investigated using statistical inference, although the approach differs by the specific method of inference.
 
@@
 If the condition is deemed unlikely to occur by chance, then there is evidence to support its complement, called the alternative hypothesis.
-The strength of evidence can be quantified, leading to decisions about what interpretation or decision to make regarding the results of the method.
+The strength of evidence can be quantified through summaries such as p-values and Bayes factors [@held2018], leading to decisions about what interpretation or decision to make regarding the results of the method.
 
@@
 For example, Q-Q plots are a common diagnostic tool for checking normality assumptions [@loy2016].
-While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist, as long as a reference model is obtainable for null data generation.
+While methods like the Shapiro-Wilk test [@shapiro1965] exist for testing normality, visual inference can be applied when assumptions are violated or tests do not exist.
 
@@
 
-In the early development of Frequentist statistics, philosophical debates between R.
+During the early development of Frequentist statistics, philosophical debates between R.
 A.
@@
 #| label: fig-freq-eval-methods
-#| fig-cap: "Frequentist evaluation methods using test statistics and confidence intervals under a null hypothesis that uses a standard normal distribution.
-Panel (a) shows the area under a standard normal for the Z-score of the observed sample mean.
-Panel (b) provides 95% confidence intervals for Z-scores of the sample mean across multiple samples."
+#| fig-cap: "Frequentist evaluation methods using test statistics and confidence intervals under a null hypothesis that uses a standard normal distribution with a sample size of 100.
+Panel (a) shows the area under the sampling distribution for where $Z=\\frac{\\bar{X}}{1/\\sqrt{100}}$ would reject the null hypothesis.
+Panel (b) provides 95% confidence intervals for sample means across multiple samples."
 #| fig-subcap: 
@@
 #| out-width: 90%
+#| fig-height: 4
+#| layout-ncol: 2
 
@@
 ggplot() + 
-  stat_function(fun = dnorm) + 
-  stat_function(fun = dnorm, xlim = c(1.5,4), 
+  stat_function(fun = \(x)(dnorm(x, sd = 1/sqrt(100))),
+                xlim = c(-4/sqrt(100), 4/sqrt(100))) + 
+  stat_function(fun = \(x)(dnorm(x, sd = 1/sqrt(100))),
+                xlim = c(qnorm(0.975, sd = 1/sqrt(100)), 4/sqrt(100)),
+                geom = 'area',
+                fill = 'red', color = 'black') +
+  stat_function(fun = \(x)(dnorm(x, sd = 1/sqrt(100))),
+                xlim = c(-qnorm(0.975, sd = 1/sqrt(100)), -4/sqrt(100)),
                 geom = 'area',
@@
   labs(x = 'Z', y = expression(phi(Z))) + 
-  scale_x_continuous(limits = c(-3, 3)) + 
+  scale_x_continuous(limits = c(-4/sqrt(100), 4/sqrt(100))) +
   theme(aspect.ratio = 1)
@@
   geom_vline(xintercept = 0, color = 'red', linetype = 'dashed') + 
-  theme(aspect.ratio = 1)
+  theme(aspect.ratio = 1, legend.position = 'none')
 ```
@@
 #| label: fig-bayes-binom
-#| fig-cap: "Caption"
-#| fig-subcap: 
-#| - "a: likelihoods"
-#| - "b: data simulated from models"
-#| layout-ncol: 2
+#| fig-cap: "Posterior distributions for probability of success of a Binomial distriubtion using Beta and uniform prior distributions.
+Data was generated so that 13 out of 20 trials were successful with a true probability of 0.7, which is denoted by the red line.
+"
+#| fig-height: 4
+#| fig-align: center
 #| out-width: 90%
@@
   scale_linetype_manual(values = prior) + 
-  labs(x = 'p', y = 'Posterior density', linetype = 'Prior')
+  labs(x = 'p', y = 'Posterior density', linetype = 'Prior') + 
+  theme(aspect.ratio = 1/2)
 
@@
   mutate(season = paste0(season-1, '-', season)) %>% 
-  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. 0 represents losses and 1 represents wins.')
+  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. 0 represents losses and 1 represents wins.', align = 'c')
 ```
@@
 #| fig-width: 6
-#| fig-cap: ""
+#| fig-cap: "Observed proportion of wins for the Minnesota Wild."
 
@@
               values_from = n, values_fill = 0) %>% 
-  mutate(propwin = result1 / (result0+result1)) %>% 
+  mutate(propwin = result1 / (result0+result1),
+         season = map_chr(season, \(x)(paste0(season-1, '-', season)))) %>% 
   filter(!(result0 == 0 & result1 == 0)) %>% 
@@
 The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are less likely to win games for larger point differentials than for smaller point differentials.
+In other words,
 
@@
 #| label: tbl-coef-compare
+
+freq.coef <- summary(mod.freq)$coefficient[,1:2] %>% 
+  as.data.frame() %>% 
+  rownames_to_column() %>% 
+  mutate(Frequentist = map2_chr(Estimate, `Std. Error`, \(x,y)(glue::glue('{round(x,2)} ({round(y, 2)})'))))
+
+bayes.u.coef <- fixef(bayes.unif)[,1:2] %>% 
+  as.data.frame() %>% 
+  rownames_to_column() %>% 
+  mutate(Bayes.unif = map2_chr(Estimate, Est.Error, \(x,y)(glue::glue('{round(x,2)} ({round(y, 2)})'))))
+
+bayes.n.coef <- fixef(bayes.priors)[,1:2] %>% 
+  as.data.frame() %>% 
+  rownames_to_column() %>% 
+  mutate(Bayes.normal = map2_chr(Estimate, Est.Error, \(x,y)(glue::glue('{round(x,2)} ({round(y, 2)})'))))
+
 data.frame(
-  Frequentist = coef(mod.freq),
-  `Flat priors` = fixef(bayes.unif)[,1],
-  `Informative priors` = fixef(bayes.priors)[,1]
+  Term = freq.coef$rowname,
+  Freq = freq.coef$Frequentist,
+  `Uniform priors` = bayes.u.coef$Bayes.unif,
+  `Informative priors` = bayes.n.coef$Bayes.normal
 ) %>% 
-  rownames_to_column(var = 'Term') %>% 
-  knitr::kable(digits = 3, caption = 'Coefficients of Frequentist and Bayesian methods for GLMs fit to the Minnesota Wild data.')
+  # rownames_to_column(var = 'Term') %>% 
+  knitr::kable(digits = 3, caption = 'Coefficients and standard errors of Frequentist and Bayesian methods for GLMs fit to the Minnesota Wild data.', col.names = c('Term', 'Frequentist GLM', 'Bayes GLM (Uniform Priors)', 'Bayes GLM (Normal Priors)'), align = 'c')
 

### 2025-03-04 paper-vis-inference/paper.qmd
Example revisions

@@
 
-Many modern statistical methods incorporate some concept of the hypothesis test [@stroup], which begin with an underlying assumption of the data generating function.
-In practice, this function is rarely known, and thus is approximated by a reasonable probability function or by attempting different models.
-For example, a generalized linear model can assume a Poisson or Negative Binomial distribution for data that has unrestricted counts, in contrast to the Binomial distribution where the total number of counts is known.
-
 When data is collected, a hypothesis test can be formulated [@eq-freq-ht].
@@
 Test statistics are formed so that their sampling distribution are known under the null hypothesis.
-For example, if the null hypothesis assume the random sample is from a normal distribution with mean $\mu$ and variance $\sigma^2$, then the sampling distribution of the sample mean $\bar{X}=\frac{1}{N}\sum_{i=1}^{N}x_i$ can be rewritten as $Z=\frac{\bar{X}-\mu}{\sigma/\sqrt{N}}$, which follows a standard normal distribution.
+For example, if the null hypothesis assume the random sample is from a normal distribution with mean $\mu$ and variance $\sigma^2$, then the sampling distribution of the sample mean, $\bar{X}=\frac{1}{N}\sum_{i=1}^{N}x_i$, can be rewritten as $Z=\frac{\bar{X}-\mu}{\sigma/\sqrt{N}}$, which follows a standard normal distribution.
 Given the data, a probability can be calculated for observing the test statistic under the null hypothesis, which is called a p-value (@fig-freq-eval-methods-1).
@@
 Since the data is random, the interpretation of Frequentist confidence intervals is that $C$ percent of confidence intervals calculated from repeated sampling of the population will contain the true parameter value $C$ percent of the time (@fig-freq-eval-methods-2).
-Evaluating hypotheses with confidence intervals simply involve checking if the parameter value set in the null hypothesis falls within the interval.
+Evaluating hypotheses with confidence intervals simply involve checking if the parameter value in the null hypothesis falls within the interval.
 If the value does not fall in the interval, then there is evidence to reject the null hypothesis.
@@
 The observed number of successes was 13 out of 20.
-Using an uninformative prior, the posterior distribution becomes @eq-binom-unif, which is proportional to a Beta distribution with $\alpha=x+1$ and $\beta=n-x+1$.
+Using an uninformative prior, the posterior distribution becomes @eq-binom-unif, which is proportional to a Beta distribution with $\alpha=13+1$ and $\beta=20-13+1$.
 
@@
 \pi(p|X))  \propto p^{x}(1-p)^{n-x}\times p^{\alpha-1}(1-p)^{\beta-1} \propto Beta(x+\alpha+1,n-x+\beta+1)
-$$
+$$ {#eq-binom-beta}
 
 Letting $\alpha=9$ and $\beta=3.857$, the prior distribution has a mean of 0.7 and a variance of 0.01515.
-The posterior distribution becomes $\text{Beta}(x+10,n-x+4.857)$, which shifts the posterior closer to 0.7 than when using the uniform prior (@fig-bayes-binom).
+The posterior distribution becomes $\text{Beta}(13+10,20-13+4.857)$, which shifts the posterior closer to 0.7 than when using the uniform prior (@fig-bayes-binom).
 
-Similar to Frequentist methods, intervals of the parameters can be created for the posterior distribution, called credible intervals [@acree2021].
-These intervals are calculated simply by encompassing a desired probability from the posterior distribution.
+Similar to Frequentist methods, intervals of the parameters can be created for the posterior distribution, which are called credible intervals [@acree2021].
+These intervals are calculated simply by determining limits for a desired probability from the posterior distribution.
 The result is that interpretations are directly about the parameter and not repeated sampling of the data.
@@
 When information is not available or there is a weak belief in prior information, a uniform prior or a prior with a large variance can be used.
-However, the choice of models can be evaluated against each other [@held2018] or averaged together [@hoeting1999; @hinne2020]
+However, the choice of models can be evaluated against each other through the Bayes factor [@held2018] or averaged together [@hoeting1999; @hinne2020]
 
@@
 Defining the target plot and null plots are essential steps in designing a lineup study.
-The target plot is chosen so that it reflects a specific condition.
-In contrast, the null plots are constructed to reflect conditions that the target graph do not satisfy.
+The target plot is chosen so that it reflects a specific condition, often involving real data.
+In contrast, the null plots are constructed to reflect conditions that the target graph do not satisfy, or simulated data.
 For example, linear models often have the assumption that residuals are independently and identically distributed with mean zero and a constant variance [@majumder2013].
@@
 The lineup protocol tests the null hypothesis that the target graph cannot be identified among graphs generated with a null model against the alternative hypothesis that the target graph can be identified among graphs generated with a null model.
-Assuming that a total of $m-1$ null plots, $1$ target plot, and $n$ participants, a Binomial distribution can be used to test the hypothesis that the target plot is chosen correctly:
+Assuming that a total of $m-1$ null plots, $1$ target plot, and $n$ participants, a Binomial distribution can be used to test the hypothesis that the target plot is chosen correctly where $\pi$ is the probability of choosing the target plot :
 
@@
 A different method of evaluation is presented by @vanderplas2021, where plot selection dependencies can be modeled with a Dirichlet-multinomial distribution.
+This accounts for various features in the plots and the strength of their signals.
 
@@
 In contrast, Bayesian inference interprets the population parameter as a random variable to quantify uncertainty.
-For the case of smallmouth bass in Lake Pepin, prior information about average weights obtained from historical records can be combined with current data to formulate the posterior distribution to update beliefs on average weights.
+For the case of smallmouth bass in Lake Pepin, prior information about average weights obtained from historical records can be combined with current data to formulate the posterior distribution to update beliefs on average weights for smallmouth bass in the lake.
 
@@
 A useful tool for both Frequentist and Bayesian inferences is a range of plausible values for the population parameter [@pek].
-For Freqentists, this range of values is referred to as a confidence interval [@neyman1937].
+For Freqentists, this range of values is the confidence interval [@neyman1937].
 The interpretations of confidence intervals are about how often repeated sampling of the population would result in the interval covering the true population parameter.
@@
 Of course, some testing procedures in visual inference rely on previous knowledge that not all null plots are created equally, and thus different signal strengths within the null plots can be incorporated in statistical significance calculations [@vanderplas2021].
-This approach is not directly a Bayesian inference, but can lead into a Bayesian approach by using the relationship between the multinomial and Dirichlet distributions.
+Although this approach is not directly a Bayesian inference, but can lead into a Bayesian approach by using the relationship between the multinomial and Dirichlet distributions.
 
@@
 
-wild <- read_csv('wild23-25.csv', show_col_types = F) %>% 
+wild <- read_csv('wild-gamelog.csv', show_col_types = F) %>% 
   mutate(pt_diff = abs(gf-ga))

### 2025-03-04 paper-vis-inference/paper.qmd
All material for submission

@@
 author: Tyler Wiederich
-date: "`r format(Sys.time(), '%B %d, %Y')`"
-format: wordcount-pdf
+date: "2025-03-01"
+date-format: long
+format: pdf
 bibliography: references.bib
@@
   warning: false
-  include: false
 ---
@@
 
-The role of the population parameter is a key differentiation between Frequentist and Bayesian methodologies [@pek].
+The role of the population parameter is a key differentiation between Frequentist and Bayesian methodologies [@pek2020].
 The former assumes the parameter to be a fixed value for its target population [@neyman1977], whereas the latter assumes the parameter is a random variable drawn from a probability distribution [@gelman2013].
@@
 #| out-width: 90%
-#| fig-height: 4
+#| fig-height: 3
 #| layout-ncol: 2
@@
 
-Unlike Frequentist methodology, there are no p-values in Bayesian hypothesis testing, at least in the sense of the Frequentist definition for a p-value [@gelman2013; @held2018] .
+Unlike Frequentist methodology, there are no p-values in Bayesian hypothesis testing, at least in the sense of the Frequentist definition for a p-value [@gelman2013; @held2018].
 Instead, the Bayes factor can be used as evidence for competing hypotheses.
@@
 #| fig-align: center
-#| out-width: 90%
 
@@
 
-The differences between Frequentist and Bayesian inference mainly arise from how population parameters are defined [@pek; @acree2021].
+The differences between Frequentist and Bayesian inference mainly arise from how population parameters are defined [@pek2020; @acree2021].
 Frequentist inference assumes that a parameter for a well-defined population at a given time is fixed and unknown, where a function of the data is the random variable.
-In some cases, this is a reasonable assumption, such as estimating the average weight of smallmouth bass in Lake Pepin, Minnesota \[[www.dnr.state.mn.us/lakefind/](https://www.dnr.state.mn.us/lakefind/showreport.html?downum=25000100)\](<https://www.dnr.state.mn.us/lakefind/showreport.html?downum=25000100>).
+In some cases, this is a reasonable assumption, such as estimating the average weight of smallmouth bass in Lake Pepin, Minnesota ([www.dnr.state.mn.us/lakefind](https://www.dnr.state.mn.us/lakefind/showreport.html?downum=25000100)).
 Here, there is a true average weight for all smallmouth bass in the lake, but the value can only be estimated.
@@
 
-A useful tool for both Frequentist and Bayesian inferences is a range of plausible values for the population parameter [@pek].
+A useful tool for both Frequentist and Bayesian inferences is a range of plausible values for the population parameter [@pek2020].
 For Freqentists, this range of values is the confidence interval [@neyman1937].
@@
 #| - "Credible interval"
+#| layout-ncol: 2
 
@@
   geom_vline(xintercept = 0, color = 'red', linetype = 'dashed') + 
-  theme(aspect.ratio = 1)
+  theme(aspect.ratio = 1, legend.position = 'none')
 
@@
 Suppose that the head coach wishes to know if the Wild are playing with an advantage playing on home-ice, meaning that they win more games at home than they do for away games.
-Data is shown in @tbl-wild.
+Data for location and results is shown in @tbl-wild.
 
@@
 wild <- read_csv('wild-gamelog.csv', show_col_types = F) %>% 
-  mutate(pt_diff = abs(gf-ga))
+  mutate(result = factor(score_rslt, levels = c('1'='W','0'='L'))) %>% 
+  select(season, location, result, gf = score_gf, starts_with('team')) %>% 
+  rename_with(.cols = starts_with('team'), 
+              .fn = str_remove, pattern = 'team_')
+
 wild %>% 
@@
   count() %>% 
-  pivot_wider(names_from = location, values_from = n) %>% 
-  mutate(season = paste0(season-1, '-', season)) %>% 
-  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. 0 represents losses and 1 represents wins.', align = 'c')
+  pivot_wider(names_from = result, values_from = n) %>% 
+  knitr::kable(caption = 'Data from the 2023-24 and 2024-25 seasons for the Minnesota Wild. L represents losses and W represents wins.', align = 'c')
 ```
@@
 The first step of any analysis should be data exploration (@fig-true-props).
-The Minnesota Wild won 48.1 percent of their home games and 63.6 percent of their away games as of February 28th, 2025.
-Immediately noticeable is that the Wild seem to perform better when not playing on home-ice.
-We also have information about point differentials that may be helpful, which is the difference in scores at the end of the game.
+The Minnesota Wild won `r round(20/(20+21),1)` percent of their home games and `r round(19/(19+22),1)` percent of their away games as of February 28th, 2025.
+Immediately noticeable is that the Wild do not appear to play with any significant home-ice advantage.
+Data was also collected for the Wild's shots on goal (SOG), penalties in minutes (PIM), power play goals (PPG), and power kill opportunities (PPO).
 
@@
 #| fig-width: 6
-#| fig-cap: "Observed proportion of wins for the Minnesota Wild."
+#| fig-cap: "Observed proportion of wins for the Minnesota Wild by game location."
 
 wild %>% 
-  group_by(season, location, result, pt_diff) %>% 
+  group_by(season, location, result) %>% 
   count() %>% 
-  pivot_wider(names_from = result, names_prefix = 'result',
-              values_from = n, values_fill = 0) %>% 
-  mutate(propwin = result1 / (result0+result1),
-         season = map_chr(season, \(x)(paste0(season-1, '-', season)))) %>% 
-  filter(!(result0 == 0 & result1 == 0)) %>% 
-  ggplot(mapping = aes(x = pt_diff, y = propwin, color = location)) + 
-  geom_line() + 
-  geom_point() + 
+  ungroup() %>% 
+  group_by(season, location) %>% 
+  mutate(prop = n / sum(n)) %>% 
+  filter(result == 'W') %>%
+  ggplot(mapping = aes(x = location, y = prop, fill = location)) + 
+  geom_bar(stat = 'identity', width = 1/2) +
   scale_y_continuous(limits = c(0,1)) + 
-  labs(x = 'Point differential',
+  labs(x = 'Location',
        y = 'Proportion of wins') +
-  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
+  scale_fill_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
   facet_wrap(~season) + 
@@
 
+@eq-freq-glm-model is the model fit to the data for the 2024-25 season, where $\beta_0$ is the intercept term, $\beta_{1i}$ is the effect for location $i$, $\beta_3$ - $\beta_5$ are the effects of their corresponding variables, and $\pi_i$ is the probability of winning for location $i$.
+
+$$
+\eta_{ijklm}=\log(\frac{\pi_{ijklm}}{(1-\pi_{ijklm})})=\beta_0+\beta_{1i}+\beta_2\cdot \text{sog}_j + \beta_3\cdot \text{pim}_k + \beta_4\cdot \text{ppg}_l + \beta_5\cdot\text{ppo}_m
+$$ {#eq-freq-glm-model}
+
 ```{r}
@@
 
+library(pander)
 options(contrasts = c("contr.sum", "contr.poly"))
 
-mod.freq <- glm(result ~ location + I(pt_diff) + I(pt_diff^2), 
-                data = dplyr::filter(wild, season == 2025),
+mod.freq <- glm(result ~ location + sog + pim + ppg + ppo, 
+                data = dplyr::filter(wild, season == '2024-25'),
      family = 'binomial')
-pander::pander(anova(mod.freq))
+pander(car::Anova(mod.freq, type = '3'),
+       caption = 'ANOVA Table for Frequentist GLM.')
 
@@
 ```{r}
-em <- emmeans(mod.freq, ~location*pt_diff, type = 'response',
-              at = list(pt_diff = 1:6))
-ggplot(data.frame(em), aes(x = factor(pt_diff), y = prob, color = location)) + 
+#| label: fig-freq-glm-preds
+#| fig-cap: "Estimated probabilities of winning for the Frequentist GLM model."
+#| fig-align: center
+
+em <- emmeans(mod.freq, ~location, type = 'response')
+ggplot(data.frame(em), aes(x = location, y = prob, color = location)) + 
   geom_point(position = position_dodge(width = 1/4)) +
@@
 
-Fitting a Frequentist generalized linear model (GLM) as shown in @eq-freq-glm reveals that the point differential is the only significant term (@tbl-anova), which means that the Wild do not seem to have a winning advantage by playing on home-ice.
-The only conclusion from the this analysis is that there is evidence that the Minnesota Wild are less likely to win games for larger point differentials than for smaller point differentials.
-In other words,
-
-$$
-\eta_{ij}=\log(\frac{\pi_{ij}}{(1-\pi_{ij})})=\beta_0+\beta_{1i}+\beta_2X_{j}+\beta_3X_{j}^2
-$$ {#eq-freq-glm}
+Fitting the Frequentist generalized linear model (GLM) reveals that power play goals (ppg) is the only significant term (@tbl-anova), which is a reasonable result given that power plays are good scoring opportunities.
+The only conclusion from the this analysis is that there no evidence that the Minnesota Wild have a competitive advantage when playing at home.
 
@@
 @fig-freq-examp-lineup shows a lineup of ten datasets produced by simulating wins and losses.
-In this lineup, probabilities are plotted along the point differential for both away and home games.
-It is difficult to identify the true dataset, which means that the Frequentist GLM appears to fit well.
+In this lineup, probabilities are plotted along the power play goals for both away and home games.
+It is difficult to identify the true dataset, which means that the Frequentist GLM appears to fit well for the location of the games.
 
@@
 wild.freq <- wild %>% 
-  filter(season == 2025) %>% 
+  filter(season == '2024-25') %>% 
+  ungroup() %>% 
   mutate(prob = predict(mod.freq, type = 'response'),
@@
   mutate(dataset = 1:n(), 
-         display = ifelse(dataset == 3, result, sim_win))
+         win = as.numeric(result) %% 2,
+         display = ifelse(dataset == 3, win, sim_win))
 
@@
 Nine of the plots show data simulated from the fitted GLM, and one plot uses the actual data.
-The target plot is plot number (81 divided by 27)."
+The target plot is plot number (10-(4+3))."
+#| fig-align: center
 
@@
   nest() %>% 
-  mutate(glm = map(data, \(x)(glm(display ~ location + pt_diff + I(pt_diff^2),
+  mutate(glm = map(data, \(x)(glm(display ~ location + sog + pim + ppg + ppo ,
                                   data = x, family = 'binomial')))) %>% 
@@
   unnest(newdata) %>% 
-  ggplot(mapping = aes(x = pt_diff, y = result, color = location)) + 
+  ggplot(mapping = aes(x = location, y = pred, color = location)) + 
   geom_point(position = position_jitter(height = 0.01, width = 0.1),
              alpha = 1/2) +
-   geom_line(aes(y = pred)) + 
-# stat_smooth(method = "glm",
-#             formula = y ~ poly(x,2, raw =T),
-#             se = F,
-#             method.args = list(family=binomial)) + xlab("Point differential") +
+   # geom_line(aes(y = pred)) + 
   scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) +
   facet_wrap(~dataset, nrow = 2) + 
+   labs(x = 'Location', y = 'Estimated Probability of Winning',
+        color = 'Location') + 
   # scale_y_continuous(limits = c(0,1)) + 
@@
 The Bayesian approach uses the `brms` package in R to fit a Bayesian GLM to the data using the Minnesota Wild win-loss record for the 2023-2024 season.
-Two models are compared, one with uninformative (flat) priors on each model coefficient, and another that uses normal priors based on a Frequentist GLM fit for the 2023-2024 season.
-The standard errors obtained from the Frequentist GLM coefficients were small, and thus multiplied by 2 for use in the Bayes GLM with normal priors to allow for more flexibility.
+Two models are compared, one with uninformative (flat) priors on each model coefficient, and another that uses normal priors based on a Frequentist GLM fit for the 2023-2024 season (@tbl-bayes-priors).
+The estimates and standard errors are overly reliant that the results from the previous season carry over to the current season, which is somewhat reasonable since the composition of the Wild is mostly the same.
+
+```{r}
+#| label: tbl-bayes-priors
+mod.prior <- glm(result ~ location + sog + pim + ppg + ppo, 
+                 data = filter(wild, season == '2023-24'), 
+                 family = 'binomial')
+
+pander(summary(mod.prior)$coefficients[,1:2],
+       caption = 'Bayesian GLM coefficients for normal priors, generated from coefficents of a Frequentist GLM for the 2023-24 season.')
+```
 
@@
 
-mod.prior <- glm(result ~ location + pt_diff + I(pt_diff^2), data = filter(wild, season == 2024))
-
-summary(mod.prior)
-
-bayes.unif <- brm(result ~ location + pt_diff + I(pt_diff^2),
-    data = filter(wild, season == 2025),
+bayes.unif <- brm(result ~ location + sog + pim + ppg + ppo,
+    data = filter(wild, season == '2024-25'),
     save_all_pars = TRUE,
@@
 
-default_prior(result ~ location + pt_diff + I(pt_diff^2),
-    data = filter(wild, season == 2025))
+# default_prior(result ~ location + sog + pim + ppg + ppo,
+#     data = filter(wild, season == '2024-25'))
 
-bayes.priors <- brm(result ~  location + pt_diff + I(pt_diff^2),
-    data = filter(wild, season == 2025),
+bayes.priors <- brm(result ~ location + sog + pim + ppg + ppo,
+    data = filter(wild, season == '2024-25'),
     prior = c(
-      set_prior("normal(-0.4379,0.2310)", class = "b", coef = "Ipt_diffE2"),
-      set_prior("normal(0.3480,0.3002)", class = "b", coef = "location1"),
-      set_prior("normal(1.8803,1.1902)", class = "b", coef = "pt_diff"),
-      set_prior("normal(-1.0285,1.2716)", class = 'Intercept')
+      set_prior("normal(-0.485,0.3003)", class = "b", coef = "location1"),
+      set_prior("normal(0.1077,0.06424)", class = "b", coef = "pim"),
+      set_prior("normal(-0.7984,0.5134)", class = "b", coef = "ppg"),
+      set_prior("normal(0.5671,0.2861)", class = "b", coef = "ppo"),
+      set_prior("normal(-0.0716,0.05113)", class = "b", coef = "sog"),
+      set_prior("normal(-0.2001,1.498)", class = 'Intercept')
     ),
@@
 set.seed(283718921)
-bayes_factor(bayes.priors, bayes.unif)
+bf <- bayes_factor(bayes.unif, bayes.priors)
+bf
 ```
 
-The Bayes GLM coeffiecients for both models are similar to the Frequentist approach (@tbl-coef-compare).
-This result is expected since the uninformative model puts more weight on the data and the informative model has relatively small standard errors for its prior distributions.
-Additionally, the coefficients obtained using informative priors have have smaller standard errors than those obtained for the uninformative priors model and the Frequentist GLM.
-
-Comparing the two Bayesian models and using the model with informative priors as the "null hypothesis", the Bayes Factor is computed to be approximately 0.42.
-Although the standard errors of the parameters for the two models differ, there is not much evidence to prefer the informative priors over the uninformative priors.
-One discrepancy between the two Bayesian models is that 95% credible interval for the uninformative model does not consider the linear point differential to be a significant term (-0.02, -4.84), whereas the informative model does (0.62, 3.33).
-In this case, it may be better to consider additional prior distributions or use model averaging to obtain a more robust conclusion [@hinne].
+The Bayes GLM coefficients for both models are similar to the Frequentist approach (@tbl-coef-compare).
+This result is expected since the uninformative model puts more weight on the data and the informative model has information from the previous season.
+Additionally, the coefficients obtained using informative priors have have smaller standard errors than those obtained for the uninformative priors model and the Frequentist GLM, and the uninformative priors model has slightly larger standard errors than the Frequentist GLM.
 
@@
 
+Comparing the two Bayesian models and using the model with informative priors as the "null hypothesis", the Bayes Factor is computed to be approximately 0.4.
+Although the standard errors of the parameters for the two models slightly differ, there is not much evidence to prefer the informative priors over the uninformative priors.
+The conclusions obtained from both Bayesian models are the same, where only power play goals is significant according to the 95% credible intervals (@fig-bayes-results).
+
+```{r}
+#| label: fig-bayes-results
+#| fig-cap: "95% credible intervals for both Bayesian models. The model with normal priors has smaller intervals, but the conclusions are the same for both models."
+#| fig-height: 3
+
+summary(bayes.unif)$fixed[,3:4] %>% 
+  rownames_to_column(var = 'term') %>% 
+  mutate(model = 'flat priors') %>% bind_rows((summary(bayes.priors)$fixed[,3:4] %>% 
+  rownames_to_column(var = 'term') %>% 
+  mutate(model = 'normal priors'))) %>% 
+  ggplot(mapping = aes(y = term, color = model)) + 
+  geom_vline(xintercept = 0, linetype = 'dashed') +
+  geom_errorbar(aes(xmin = `l-95% CI`, xmax = `u-95% CI`), 
+                position = position_dodge(width = 1/2),
+                width = 1/4) + 
+  scale_color_brewer(palette = 'Set1') + 
+  labs(x = '95% Credible Interval',
+       y = 'Model Term') + 
+  theme(aspect.ratio = 1/2)
+
+```
+
 ```{r}
@@
 set.seed(42)
-em.unif <- emmeans(bayes.unif, ~location*pt_diff,
+em.unif <- emmeans(bayes.unif, ~location + sog + pim + ppg + ppo,
                    at = list(pt_diff = 1:6), type = 'response')
-em.priors <- emmeans(bayes.priors, ~location*pt_diff,
+em.priors <- emmeans(bayes.priors, ~location + sog + pim + ppg + ppo,
                    at = list(pt_diff = 1:6), type = 'response')
 
-ggplot(data.frame(em.unif), aes(x = factor(pt_diff), y = response, color = location)) + 
+ggplot(data.frame(em.unif), aes(x = location, y = response, color = location)) + 
   geom_point(position = position_dodge(width = 1/4)) +
@@
                 width = 1/4) + 
-  labs(x = 'Point differential', y = 'Predicted Probability') + 
+  labs(x = 'Location', y = 'Predicted Probability') + 
   scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
-  theme_bw() + theme(aspect.ratio = 1) +
+  theme_bw() + theme(aspect.ratio = 1, legend.position = 'none') +
   scale_y_continuous(limits = c(0,1))
 
-ggplot(data.frame(em.priors), aes(x = factor(pt_diff), y = response, color = location)) + 
+ggplot(data.frame(em.priors), aes(x = location, y = response, color = location)) + 
   geom_point(position = position_dodge(width = 1/4)) +
@@
                 width = 1/4) + 
-  labs(x = 'Point differential', y = 'Predicted Probability') + 
+  labs(x = 'Location', y = 'Predicted Probability') + 
   scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
-  theme_bw() + theme(aspect.ratio = 1) +
+  theme_bw() + theme(aspect.ratio = 1, legend.position = 'none') +
   scale_y_continuous(limits = c(0,1))
@@
 Similar to the Frequentist case, visual inference can be used to examine model diagnostics for the Bayesian models.
-Here, it would be worth using comparing the uninformative and informative models to provide additional evidence along with the Bayes factor for which model to use since the linear term of point differential has conflicting conclusions.
+Here, it would be worth using comparing the uninformative and informative models to evaluate if either model is sufficient.
+Additionally, plotting predicted probabilities along each model term would be good candidates to use as target plots.
+Data can be simulated from the probabilities, refit with with model, and placed into a lineup study with the target plot.
+
+```{r}
+#| label: fig-bayes-vis-targets
+#| fig-cap: "Potential target plots for visual inference"
+#| fig-width: 6
+wild %>% 
+  filter(season == '2024-25') %>% 
+  mutate(pred.unif = predict(bayes.unif, type = 'response')[,1],
+         pred.norm = predict(bayes.priors, type = 'response')[,1]) %>% 
+  filter(result == 'W') %>% 
+  pivot_longer(sog:ppo, names_to = 'term') %>% 
+  rename(
+    `Uniform Priors` = pred.unif,
+    `Normal Priors` = pred.norm
+  ) %>% 
+  pivot_longer(`Uniform Priors`:`Normal Priors`, names_to = 'model', values_to = 'prob') %>% 
+  ggplot(mapping = aes(x = value, y = prob, color = location)) + 
+  labs(x = expression(theta),
+       y = 'Predicted Probability of Winning') + 
+  geom_point(alpha = 1/2) + 
+  scale_color_manual(values = c('home' = '#154734', 'away' = '#A6192E')) + 
+  facet_grid(model~term, scales = 'free') + 
+  theme(aspect.ratio = 1)
+```
+
+For the Minnesota Wild, the probability of winning games appears to be a coin flip.
+Only power play goals had a significant effect, as indicated in the Frequentist and Bayesian GLMs.
+However, it was interesting to see that the Wild do not seem to have an advantage while playing games at the Excel Energy Center.
 
@@
 The commonality of statistical methods is that uncertainty is quantified through a random variable, but defining the random variable is where the methods diverge in computation and interpretations.
-In this paper, Frequentist and Bayesian inferences were compared along with visual inference, and when each methodology should be used.
+In this paper, the methodology of Frequentist, Bayesian, and visual inferences were discussed and compared, included when each type of inference should be used.
 
@@
 This is a reasonable situation when conducting novel experiments or collecting data from new sources.
-Statements of uncertainty pertain to what happens under repeated sampling conditions, which is useful when wanting to know what conditions to expect for the next instance of data collection.
+Statements of uncertainty pertain to what happens under repeated sampling conditions, which is useful when wanting to know what conditions to expect for the next instance of similar data collection.
 Results are more easily calculated, having test statistics and confidence intervals based on sampling distributions of statistics from the parameter.
@@
 Other than model checking, this is where Bayesian analyses typically end since the goal is to make make inferences on the parameter(s).
-Bayesian methods are typically computationally demanding due to the complex posterior calculations.
+Bayesian methods are typically computationally demanding due to the complex posterior calculations
 
 Lastly, visual inference is unique in that it can function on its own or in addition to Frequentist and Bayesian methods [@loy2017; @gelman2013].
-The main goal of lineup studies is to determine if a target plot can be identified.
-If it can be distinguished, then there is evidence that humans a particular feature of the target plot.
+The main goal of lineup studies is to determine if a target plot can be identified from a set of null plots.
+If it can be distinguished, then there is evidence that humans can pick up on particular features of the target plot.
 From @fig-lineup-examp1, null plots were modeled with a gamma distribution for the y-axis variable and a normal distribution for the x-axis variable.
